---
title: "Statistical Learning Foundations"
---

Statistical learning is the study of using data to estimate structure,
make predictions, and quantify uncertainty. The course begins with a
single guiding idea: most statistical learning methods are trying to
learn something about an unknown relationship between inputs and an
outcome, or about structure inside a data set.

The methods in this course range from ordinary least squares to
resampling, regularization, classification, trees, support vector
machines, and unsupervised learning. They differ in notation and
implementation, but they share a common workflow: define the learning
problem, choose a model class, fit the model, assess performance, and
communicate what the model can and cannot support.

## Learning Objectives

- Define statistical learning and explain how it overlaps with machine
  learning.
- Distinguish supervised, unsupervised, regression, and classification
  problems.
- Separate prediction goals from inference goals.
- Write the basic supervised learning model $Y = f(X) + \epsilon$.
- Define reducible error, irreducible error, training error, and test
  error.
- Explain why model flexibility creates a trade-off between underfitting
  and overfitting.
- Carry out an initial data inspection before fitting a first model.

## Why Learn From Data?

Statistical learning is useful when a problem contains patterns that are
too complex, too noisy, or too data-dependent to be fully described by
fixed rules. Examples include predicting a patient's risk after a heart
attack, estimating future demand from economic indicators, recognizing
handwritten digits, identifying spam, recommending movies, or modeling
housing prices from neighborhood characteristics.

The phrase "machine learning" often emphasizes algorithms, automation,
scale, and prediction accuracy. The phrase "statistical learning" often
emphasizes models, uncertainty, interpretation, and principled
assessment. In practice, the boundary is blurry. A good analyst borrows
from both traditions: use flexible tools when they improve performance,
and use statistical thinking to understand what the evidence supports.

The first habit is to name the task before choosing a method. A method
that is excellent for predicting new outcomes may not give simple
coefficient interpretations. A method that gives clear interpretations
may not be the strongest predictor. The goal controls the evidence we
need.

## Learning Problems

In supervised learning, each observation includes an outcome. The model
uses predictors to estimate or predict that outcome. In unsupervised
learning, there is no designated outcome; the goal is to discover
structure among observations or variables.

| Problem type | Outcome | Typical question | Example |
|---|---|---|---|
| Regression | Quantitative | What value should we predict? | Predict sales or home value. |
| Classification | Categorical | Which class or probability should we assign? | Predict default or disease status. |
| Clustering | None | Which observations are similar? | Segment customers by behavior. |
| Dimension reduction | None | Can many variables be summarized by fewer features? | Build principal components from correlated measures. |
| Anomaly detection | None or weak labels | Which cases are unusual? | Flag suspicious transactions. |

Regression and classification are supervised tasks. Clustering and many
dimension-reduction methods are unsupervised tasks. A single applied
project can combine both: for example, cluster customers to describe
groups, then use those groups as features in a supervised prediction
model.

## A Common Notation

Suppose the data contain $n$ observations and $p$ predictors. For
observation $i$, write the predictors as

$$
\begin{align}
X_i = (X_{i1}, X_{i2}, \ldots, X_{ip})
\end{align}
$$

and the outcome as $Y_i$. In a supervised learning problem, we assume
there is some unknown relationship

$$
\begin{align}
Y = f(X) + \epsilon,
\end{align}
$$

where $f(X)$ is the systematic part of the relationship and $\epsilon$
collects everything not explained by the predictors: measurement error,
omitted variables, randomness, and unmodeled variation.

The regression function is

$$
\begin{align}
f(x) = \operatorname{E}(Y \mid X = x).
\end{align}
$$

A fitted model $\hat f$ is an estimate of $f$. A simple linear model
estimates $f$ with a line or plane. A tree, random forest, spline, or
support vector machine estimates $f$ with a more flexible shape. The
question is not which method is universally best. The question is which
method learns useful structure for the problem, data, and decision at
hand.

## Prediction, Inference, Description, and Decisions

Prediction asks for accurate values of $\hat Y$ on future or held-out
cases. Inference asks what the data suggest about the relationship
between variables. Description asks what patterns are visible in the
data. Decision support asks what action should be taken after accounting
for error costs and uncertainty.

| Goal | Main question | Typical evidence |
|---|---|---|
| Prediction | How well does the model generalize? | Validation error, test error, calibration |
| Inference | Which relationships are supported? | Coefficients, uncertainty intervals, diagnostics |
| Description | What patterns are visible? | Summaries, plots, cluster profiles |
| Decision support | What action should be taken? | Error costs, constraints, uncertainty |

These goals can overlap, but they should not be confused. A model can
predict well and still be difficult to interpret. A model can have
interpretable coefficients and still predict poorly outside the training
data. A model can describe an association and still fail to justify a
causal claim.

## Reducible and Irreducible Error

At a new point $x_0$, the prediction error for a quantitative outcome
can be written as

$$
\begin{align}
\operatorname{E}\{(Y - \hat f(x_0))^2\}
= \operatorname{E}\{(f(x_0) - \hat f(x_0))^2\}
+ \operatorname{Var}(\epsilon).
\end{align}
$$

The first term is reducible error. It may decrease if we collect better
data, engineer better predictors, choose a better model class, tune the
method more carefully, or use more observations. The second term is
irreducible error. Even if $f$ were known perfectly, two observations
with the same predictors could still have different outcomes.

This distinction keeps model assessment realistic. A model can be
improved, but not every error is a modeling failure. Some uncertainty is
part of the data-generating process.

## Flexibility, Underfitting, and Overfitting

Model flexibility describes how many shapes a method can learn. A linear
model is relatively inflexible. A high-degree polynomial, deep tree, or
large ensemble can be much more flexible.

| Model behavior | What happens | Common symptom |
|---|---|---|
| Underfitting | The model is too rigid to capture real structure. | High training error and high test error. |
| Useful fit | The model captures stable signal without chasing noise. | Low test error relative to simpler alternatives. |
| Overfitting | The model adapts to noise in the training data. | Very low training error but high test error. |

Increasing flexibility usually lowers training error. It does not always
lower test error. Test error often decreases at first as the model
captures real structure, then increases when the model begins to chase
noise. This is the practical version of the bias-variance trade-off:
rigid models can have high bias, while overly flexible models can have
high variance.

## Training Error and Test Error

Training error is computed on the same observations used to fit the
model. Test error is computed on observations not used to fit or tune
the model. Because the model has already adapted to the training data,
training error is usually too optimistic as a measure of future
performance.

For regression, a common loss is mean squared error:

$$
\begin{align}
\operatorname{MSE}
= \frac{1}{n}\sum_{i=1}^{n}(y_i - \hat y_i)^2.
\end{align}
$$

For classification, a common measure is the misclassification rate:

$$
\begin{align}
\operatorname{MCE}
= \frac{1}{n}\sum_{i=1}^{n} I(y_i \neq \hat y_i).
\end{align}
$$

Accuracy alone can be misleading when classes are rare or when different
mistakes have different costs. Later chapters will add sensitivity,
specificity, precision, ROC curves, and calibration. The important first
principle is simple: evaluate the model on data that were not used to
choose it.

## The Statistical Learning Workflow

A reproducible statistical learning analysis usually follows this
sequence:

1. State the question and identify the outcome, predictors, and unit of
   observation.
2. Inspect the data for scale, missingness, unusual values, and visible
   relationships.
3. Choose a baseline model that is simple enough to understand.
4. Split the data or use resampling to estimate out-of-sample
   performance.
5. Fit candidate models and compare them using evidence tied to the
   project goal.
6. Diagnose limitations: assumptions, extrapolation, instability,
   confounding, fairness concerns, and data quality.
7. Communicate the result in language that matches the evidence.

The baseline model is not a formality. It provides a reference point.
If a more complex method does not improve on a simple baseline, the
simple model may be the better choice.

## Opening Example: Housing Data

A first analysis should begin before modeling. In the Boston housing
data, the observational unit is a census tract, the response `medv` is
median home value, and candidate predictors include crime rate,
industrial land use, distance to employment centers, pupil-teacher
ratio, and other tract-level summaries. The data are historical and
should be used as a methods example rather than as a current policy
description.

Before fitting a model, answer five questions:

1. What is the response, and what does one row represent?
2. Which predictors are quantitative and which are categorical or
   binary?
3. Are there unusual values, impossible values, or strong skew?
4. Which pairwise relationships with the response are most visible?
5. Which baseline model is simple enough to interpret?

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: true
#| fig-align: center
library(ggplot2)
library(tidyverse)
library(reticulate)
reticulate::py_install("pandas")
reticulate::py_install("scikit-learn")
reticulate::py_install("scipy")

boston <- read.csv("../datasets/Boston.csv")
```

```{r}
#| echo: true
#| eval: true
#| fig-align: center
str(boston)
summary(boston)
```

```{r}
#| echo: true
#| eval: true
#| fig-align: center
#| fig-cap: "A simple baseline relationship between percent lower status population and median home value."
#| fig-alt: "A scatterplot of median home value against percent lower status population, with a fitted line showing a negative relationship."
#| label: fig-baseline
baseline_g <- ggplot(boston, aes(lstat, medv)) +
  geom_point(alpha = 0.65) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Percent lower status population",
    y = "Median home value",
    title = "A simple baseline relationship"
  );

baseline_g

```

```{r}
#| echo: true
#| eval: true
#| fig-align: center
#| output: asis

baseline <- lm(medv ~ lstat, data = boston)
expanded <- lm(medv ~ lstat + rm, data = boston)

# summary(baseline)
# summary(expanded)
stargazer::stargazer(baseline, expanded, type = "html",summary=TRUE,report = "vc*stp",ci=TRUE)
```

## Python

```{python}
#| echo: true
#| eval: true
#| fig-align: center

import pandas as pd
import seaborn as sns
import statsmodels.formula.api as smf
import matplotlib.pyplot as plt

boston = pd.read_csv("../datasets/Boston.csv")

print(boston.info())
print(boston.describe())

sns.regplot(data=boston, x="lstat", y="medv", scatter_kws={"alpha": 0.65})
plt.xlabel("Percent lower status population")
plt.ylabel("Median home value")
plt.title("A simple baseline relationship")
plt.show()

baseline = smf.ols("medv ~ lstat", data=boston).fit()
expanded = smf.ols("medv ~ lstat + rm + ptratio + crim", data=boston).fit()

print(baseline.summary())
print(expanded.summary())

```

:::

The purpose of this first fit is orientation. The baseline model asks
whether one visible predictor carries useful signal. The expanded model
asks whether additional predictors change prediction and interpretation.
Neither model should be treated as final until it has been checked on
held-out data and examined through diagnostics.

## First Train/Test Split

The next step is to estimate how the model performs away from the data
used to fit it.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: true
set.seed(571)

id <- sample(seq_len(nrow(boston)), size = 0.8 * nrow(boston))
train <- boston[id, ]
test <- boston[-id, ]

fit <- lm(medv ~ lstat + rm + ptratio + crim, data = train)
pred <- predict(fit, newdata = test)

rmse <- sqrt(mean((test$medv - pred)^2))
test_r2 <- 1 - sum((test$medv - pred)^2) /
  sum((test$medv - mean(test$medv))^2)

c(rmse = rmse, test_r2 = test_r2)
```

## Python

```{python}
#| echo: true
#| eval: true
from sklearn.model_selection import train_test_split
from sklearn.metrics import mean_squared_error, r2_score

train, test = train_test_split(boston, train_size=0.8, random_state=571)

fit = smf.ols("medv ~ lstat + rm + ptratio + crim", data=train).fit()
pred = fit.predict(test)

rmse = mean_squared_error(test["medv"], pred) ** 0.5
test_r2 = r2_score(test["medv"], pred)

print({"rmse": round(rmse,4), "test_r2": round(test_r2,4)})
```

:::

Residual standard error describes typical residual size in the training
data. Test RMSE describes typical prediction error on observations not
used to estimate the coefficients. Both can be useful, but they answer
different questions.

## Assumptions as Modeling Questions

Ordinary least squares relies on assumptions that should be treated as
diagnostic questions rather than as boxes to check mechanically:

| Assumption | Practical question |
|---|---|
| Linearity | Is the conditional mean roughly linear in the predictors used? |
| Independent errors | Are observations collected in a way that creates dependence? |
| Constant variance | Does residual spread change with fitted values or predictors? |
| Approximate normality | Are interval estimates relying on a poor error approximation? |
| No extreme leverage | Are a few observations controlling the fitted relationship? |

The assumptions matter because they affect interpretation,
uncertainty, and prediction. When a diagnostic reveals a problem, the
next step may be a transformation, an interaction, a nonlinear term, a
different model class, or a narrower claim.

## Closing Checklist

By the end of a first statistical learning analysis, you should be able
to say:

- The unit of observation is clear.
- The response and predictors are defined.
- The learning task is regression, classification, unsupervised
  learning, or a combination.
- The goal is prediction, inference, description, decision support, or a
  named mixture of these.
- A simple baseline model has been fit.
- Performance has been assessed on held-out data or with a resampling
  plan.
- The main limitations of the model are stated plainly.
