---
title: "Linear Regression"
---

Linear regression is the first complete statistical learning workflow:
specification, estimation, interpretation, diagnostics, uncertainty,
and prediction. It is simple enough to study carefully and rich enough
to introduce most habits used throughout the course.

## Learning Objectives

- Write simple and multiple linear regression models.
- Interpret coefficients for continuous variables, categorical
  variables, and interactions.
- Explain least squares as loss minimization.
- Evaluate a regression model with residual diagnostics and held-out
  prediction error.
- Translate a fitted model into a clear substantive statement.

## Least Squares Regression

The multiple linear regression model has the form

$$
\begin{align}
Y_i = \beta_0 + \beta_1X_{i1} + \cdots + \beta_pX_{ip}
+ \epsilon_i.
\end{align}
$$

Ordinary least squares chooses coefficients that minimize residual sum
of squares:

$$
\begin{align}
\hat{\beta} =
\arg\min_{\beta}
\sum_{i=1}^{n}(y_i - x_i^\top\beta)^2.
\end{align}
$$

The fitted value is $\hat y_i = x_i^\top\hat\beta$, and the residual is
$e_i = y_i - \hat y_i$. A coefficient estimates the expected change in
the response for a one-unit increase in that predictor, holding the
other predictors in the model fixed. That "holding fixed" phrase is
essential: coefficient meaning depends on the full model specification.

## Advertising Model Sequence

The advertising data provide a compact way to move from simple
regression to multiple regression. Suppose `sales` is the response and
`TV`, `radio`, and `newspaper` are advertising budgets.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: true
#| fig-align: center
set.seed(571)
advertising <- data.frame(
  TV = runif(200, 0, 300),
  radio = runif(200, 0, 50),
  newspaper = runif(200, 0, 100)
)
advertising$sales <- 3 +
  0.045 * advertising$TV +
  0.19 * advertising$radio +
  0.002 * advertising$TV * advertising$radio +
  rnorm(200, sd = 2)

tv_fit <- lm(sales ~ TV, data = advertising)
media_fit <- lm(sales ~ TV + radio + newspaper, data = advertising)
interaction_fit <- lm(sales ~ TV * radio + newspaper, data = advertising)

summary(tv_fit)
summary(media_fit)
summary(interaction_fit)
```

## Python

```{python}
#| echo: true
#| eval: true
#| fig-align: center
import numpy as np
import pandas as pd
import statsmodels.formula.api as smf

rng = np.random.default_rng(571)
advertising = pd.DataFrame({
    "TV": rng.uniform(0, 300, 200),
    "radio": rng.uniform(0, 50, 200),
    "newspaper": rng.uniform(0, 100, 200),
})
advertising["sales"] = (
    3
    + 0.045 * advertising["TV"]
    + 0.19 * advertising["radio"]
    + 0.002 * advertising["TV"] * advertising["radio"]
    + rng.normal(0, 2, 200)
)

tv_fit = smf.ols("sales ~ TV", data=advertising).fit()
media_fit = smf.ols("sales ~ TV + radio + newspaper", data=advertising).fit()
interaction_fit = smf.ols(
    "sales ~ TV * radio + newspaper",
    data=advertising
).fit()

print(tv_fit.summary())
print(media_fit.summary())
print(interaction_fit.summary())
```

:::

The simple model asks whether `TV` budget alone predicts sales. The
multiple regression model asks whether each medium explains variation
after accounting for the others. The interaction model asks whether the
effect of one medium changes as another medium changes.

## Train/Test Evaluation

Regression teaching often begins with coefficient tables, but
prediction requires held-out evaluation. Split before comparing
candidate models:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: true
#| fig-align: center
set.seed(571)
id <- sample(seq_len(nrow(advertising)), size = 0.8 * nrow(advertising))
train <- advertising[id, ]
test <- advertising[-id, ]

fit <- lm(sales ~ TV + radio + newspaper, data = train)
pred <- predict(fit, newdata = test)

rmse <- sqrt(mean((test$sales - pred)^2))
test_r2 <- 1 - sum((test$sales - pred)^2) /
  sum((test$sales - mean(test$sales))^2)
```

## Python

```{python}
#| echo: true
#| eval: true
#| fig-align: center
from sklearn.model_selection import train_test_split
from sklearn.metrics import mean_squared_error, r2_score
import statsmodels.formula.api as smf

train, test = train_test_split(
    advertising,
    train_size=0.8,
    random_state=571
)

fit = smf.ols("sales ~ TV + radio + newspaper", data=train).fit()
pred = fit.predict(test)

rmse = mean_squared_error(test["sales"], pred) ** 0.5
test_r2 = r2_score(test["sales"], pred)

print({"rmse": rmse, "test_r2": test_r2})
```

:::

Residual standard error describes typical residual size in the training
data. Test RMSE describes typical prediction error on observations not
used to estimate the coefficients. Both can be useful, but they answer
different questions.

## Interactions and Marginal Effects

An interaction lets the effect of one predictor depend on another:

$$
\operatorname{E}(Y \mid X,Z)
= \beta_0 + \beta_1X + \beta_2Z + \beta_3XZ.
$$

The marginal effect of $X$ is

$$
\frac{\partial \operatorname{E}(Y \mid X,Z)}{\partial X}
= \beta_1 + \beta_3Z.
$$

After an interaction is included, the main effect of $X$ is the effect
when $Z=0$. If $Z=0$ is not meaningful, centering $Z$ can make the main
effect easier to interpret.

## Diagnostics and Influence

Diagnostics convert assumptions into visible questions:

| Diagnostic | Question |
|---|---|
| Residuals vs. fitted values | Is the mean structure missing curvature or interactions? |
| Scale-location plot | Does residual spread change across fitted values? |
| Normal Q-Q plot | Are uncertainty summaries sensitive to tail behavior? |
| Leverage | Are some predictor combinations unusual? |
| Cook's distance | Are individual observations strongly changing the fit? |
| Correlation or VIF | Are coefficients unstable because predictors overlap? |

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: true
#| fig-align: center
par(mfrow = c(2, 2))
plot(fit)
```

## Python

```{python}
#| echo: true
#| eval: true
#| fig-align: center
import matplotlib.pyplot as plt
import scipy.stats as stats

residuals = fit.resid
fitted = fit.fittedvalues

fig, axes = plt.subplots(1, 2, figsize=(10, 4))
axes[0].scatter(fitted, residuals)
axes[0].axhline(0, color="black", linewidth=1)
axes[0].set_xlabel("Fitted values")
axes[0].set_ylabel("Residuals")
stats.probplot(residuals, dist="norm", plot=axes[1])
plt.tight_layout()
plt.show()
```

:::

The goal is not to make every plot perfect. The goal is to notice
model-data disagreement and decide whether it changes the claim.

## Coefficient Interpretation

Good interpretation follows a stable template:

1. name the response scale,
2. name the predictor and unit change,
3. state the direction and magnitude,
4. state what is held fixed,
5. avoid causal language unless the design supports it.

For example: holding radio and newspaper budgets fixed, an additional
unit of TV advertising budget is associated with an estimated increase
of $\hat\beta_1$ sales units. This is an association from an
observational regression, not by itself proof that changing TV spending
would cause the same change in sales.
