---
title: "Interactions and Marginal Effects"
---

Interactions are the first place where a regression model stops being a
collection of independent additive effects. They let the relationship
between one predictor and the response depend on another predictor.

## Learning Objectives

- Write a regression model with an interaction term.
- Interpret main effects when an interaction is present.
- Compute marginal effects for binary-continuous and
  continuous-continuous interactions.
- Use plots to communicate conditional relationships.
- Explain why interactions should be motivated by the question and
  checked against the data.

## Interaction Model

For two predictors $X$ and $Z$, an interaction model is

$$
\operatorname{E}(Y \mid X,Z)
= \beta_0 + \beta_1X + \beta_2Z + \beta_3XZ.
$$

The coefficient $\beta_3$ says how the slope of $X$ changes as $Z$
changes. The marginal effect of $X$ is

$$
\frac{\partial \operatorname{E}(Y \mid X,Z)}{\partial X}
= \beta_1 + \beta_3Z.
$$

The main effect $\beta_1$ is therefore the effect of $X$ when $Z=0$.
If $Z=0$ is not substantively meaningful, centering $Z$ makes the main
effect easier to explain.

## Binary by Continuous Interaction

Suppose the return to advertising differs by whether a campaign is
digital-first. The model includes a continuous predictor, a binary
indicator, and their product.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: true
#| fig-align: center
set.seed(571)
n <- 250
campaign <- data.frame(
  spend = runif(n, 10, 100),
  digital = rbinom(n, size = 1, prob = 0.45)
)
campaign$sales <- 20 + 0.55 * campaign$spend +
  7 * campaign$digital +
  0.35 * campaign$spend * campaign$digital +
  rnorm(n, sd = 8)

fit_interaction <- lm(sales ~ spend * digital, data = campaign)
summary(fit_interaction)
```

## Python

```{python}
#| echo: true
#| eval: true
#| fig-align: center
import numpy as np
import pandas as pd
import statsmodels.formula.api as smf

rng = np.random.default_rng(571)
n = 250
campaign = pd.DataFrame({
    "spend": rng.uniform(10, 100, n),
    "digital": rng.binomial(1, 0.45, n)
})
campaign["sales"] = (
    20 + 0.55 * campaign["spend"] +
    7 * campaign["digital"] +
    0.35 * campaign["spend"] * campaign["digital"] +
    rng.normal(0, 8, n)
)

fit_interaction = smf.ols("sales ~ spend * digital", data=campaign).fit()
print(fit_interaction.summary())
```

:::

The slope for non-digital campaigns is $\beta_1$. The slope for
digital-first campaigns is $\beta_1+\beta_3$. The intercept difference
is not the whole story because the gap between groups changes with
spending.

## Continuous by Continuous Interaction

When both predictors are continuous, the effect of each predictor is a
function of the other. A coefficient table alone is rarely enough. Use
predicted values over a grid of meaningful values.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: true
#| fig-align: center
new_grid <- expand.grid(
  spend = seq(10, 100, length.out = 50),
  digital = c(0, 1)
)
new_grid$predicted_sales <- predict(fit_interaction, newdata = new_grid)
head(new_grid)
```

## Python

```{python}
#| echo: true
#| eval: true
#| fig-align: center
new_grid = pd.DataFrame({
    "spend": np.tile(np.linspace(10, 100, 50), 2),
    "digital": np.repeat([0, 1], 50)
})
new_grid["predicted_sales"] = fit_interaction.predict(new_grid)
print(new_grid.head())
```

:::

This prediction-grid approach also works for interactions between
income and age, education and experience, price and promotion, or any
other pair where the substantive effect is conditional.

## Standard Errors for Marginal Effects

For a continuous-continuous interaction, the marginal effect
$\beta_1+\beta_3Z$ has uncertainty because it depends on two estimated
coefficients. Its variance is

$$
\operatorname{Var}(\hat\beta_1 + z\hat\beta_3)
= \operatorname{Var}(\hat\beta_1)
+ z^2\operatorname{Var}(\hat\beta_3)
+ 2z\operatorname{Cov}(\hat\beta_1,\hat\beta_3).
$$

The uncertainty can be narrow in some regions of $Z$ and wide in
others. This is why marginal-effect plots are often more honest than a
single starred interaction coefficient.
