---
title: "Model Assessment and Resampling"
---

Resampling methods turn model fitting into model evaluation. Instead of
trusting one fitted model or one arbitrary train/test split, we
repeatedly refit a model on related versions of the data and study how
its predictions or estimates change. This chapter uses resampling to
answer two practical questions: how well will a model predict new data,
and how much uncertainty is attached to a fitted statistic or model
coefficient?

## Learning Objectives

- Distinguish training error, validation error, test error, and
  cross-validation error.
- Explain why a single validation split can be unstable.
- Implement leave-one-out and k-fold cross-validation.
- Use resampling to compare candidate regression models.
- Use the bootstrap to estimate uncertainty in statistics and model
  coefficients.

## Why Training Error Is Not Enough

A model is useful only if it generalizes. Training error measures how
well the fitted model describes data it has already seen. Because the
model was chosen to fit those observations, training error is usually
optimistic. The more flexible the model, the easier it is to drive
training error down without improving future predictions.

The validation-set approach addresses this optimism by splitting the
data into a training set and a held-out validation or test set. The
model is fit only on the training set, and prediction error is computed
on observations not used during fitting.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(ISLR)
library(boot)
library(ggplot2)

auto <- na.omit(ISLR::Auto)

set.seed(571)
train_id <- sample(seq_len(nrow(auto)), size = 0.6 * nrow(auto))
train <- auto[train_id, ]
test <- auto[-train_id, ]

fit <- lm(mpg ~ horsepower, data = train)
pred <- predict(fit, newdata = test)
mean((test$mpg - pred)^2)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
import pandas as pd
import statsmodels.formula.api as smf
from sklearn.model_selection import train_test_split
from sklearn.metrics import mean_squared_error

rng = np.random.default_rng(571)
auto = pd.DataFrame({
    "horsepower": rng.uniform(45, 230, 392)
})
auto["mpg"] = (
    48 - 0.18 * auto["horsepower"] +
    0.00035 * auto["horsepower"] ** 2 +
    rng.normal(0, 3, len(auto))
)

train, test = train_test_split(auto, train_size=0.6, random_state=571)
fit = smf.ols("mpg ~ horsepower", data=train).fit()
pred = fit.predict(test)
print(mean_squared_error(test["mpg"], pred))
```

:::

The validation-set estimate is easy to understand, but it depends on
which observations happen to fall in the training set. If we change the
random split, the estimated test error can change as well.

## Polynomial Complexity in the `Auto` Data

The `Auto` data show a curved relationship between horsepower and gas
mileage. A straight line is too rigid, but an extremely high-degree
polynomial may fit noise. Resampling gives a disciplined way to compare
model flexibility.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
ggplot(auto, aes(horsepower, mpg)) +
  geom_point(alpha = 0.65) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(
    method = "lm",
    formula = y ~ poly(x, 2),
    se = FALSE,
    color = "#2f5f7e"
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ poly(x, 4),
    se = FALSE,
    color = "#7a3035"
  )
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import matplotlib.pyplot as plt
import seaborn as sns

sns.regplot(data=auto, x="horsepower", y="mpg", order=1, scatter_kws={"alpha": 0.65})
sns.regplot(data=auto, x="horsepower", y="mpg", order=2, scatter=False, color="#2f5f7e")
sns.regplot(data=auto, x="horsepower", y="mpg", order=4, scatter=False, color="#7a3035")
plt.show()
```

:::

For one split, compute validation error for polynomial degrees 1
through 10:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
set.seed(571)
train_id <- sample(seq_len(nrow(auto)), size = 0.6 * nrow(auto))
train <- auto[train_id, ]
test <- auto[-train_id, ]

validation_error <- sapply(1:10, function(degree) {
  fit <- lm(mpg ~ poly(horsepower, degree), data = train)
  pred <- predict(fit, newdata = test)
  mean((test$mpg - pred)^2)
})

data.frame(degree = 1:10, mse = validation_error)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.linear_model import LinearRegression
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import PolynomialFeatures

validation_error = []

for degree in range(1, 11):
    model = make_pipeline(
        PolynomialFeatures(degree, include_bias=False),
        LinearRegression()
    )
    model.fit(train[["horsepower"]], train["mpg"])
    pred = model.predict(test[["horsepower"]])
    validation_error.append(mean_squared_error(test["mpg"], pred))

print(pd.DataFrame({"degree": range(1, 11), "mse": validation_error}))
```

:::

The usual pattern is a sharp improvement from degree 1 to degree 2 and
little consistent gain from higher degrees. The important lesson is not
the exact best degree for one split; it is that the answer can vary
when the split changes.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
split_errors <- do.call(rbind, lapply(1:10, function(seed) {
  set.seed(seed)
  train_id <- sample(seq_len(nrow(auto)), size = 0.6 * nrow(auto))
  train <- auto[train_id, ]
  test <- auto[-train_id, ]

  data.frame(
    split = seed,
    degree = 1:10,
    mse = sapply(1:10, function(degree) {
      fit <- lm(mpg ~ poly(horsepower, degree), data = train)
      pred <- predict(fit, newdata = test)
      mean((test$mpg - pred)^2)
    })
  )
}))

ggplot(split_errors, aes(degree, mse, group = split)) +
  geom_line(alpha = 0.45) +
  geom_point(alpha = 0.45)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
split_records = []

for seed in range(1, 11):
    train_i, test_i = train_test_split(auto, train_size=0.6, random_state=seed)
    for degree in range(1, 11):
        model = make_pipeline(
            PolynomialFeatures(degree, include_bias=False),
            LinearRegression()
        )
        model.fit(train_i[["horsepower"]], train_i["mpg"])
        pred = model.predict(test_i[["horsepower"]])
        split_records.append({
            "split": seed,
            "degree": degree,
            "mse": mean_squared_error(test_i["mpg"], pred)
        })

split_errors = pd.DataFrame(split_records)
sns.lineplot(data=split_errors, x="degree", y="mse", units="split", estimator=None, alpha=0.45)
plt.show()
```

:::

Repeated splits usually agree that a quadratic model is much better
than a line, but they need not agree on every detail. This instability
motivates cross-validation.

## Leave-One-Out Cross-Validation

Leave-one-out cross-validation (LOOCV) uses each observation once as a
validation point. For observation $i$, fit the model on all observations
except $i$, predict the held-out response, and compute

$$
MSE_i = (y_i - \hat y_i)^2.
$$

The LOOCV estimate averages these $n$ errors:

$$
CV_{(n)} = \frac{1}{n}\sum_{i=1}^{n} MSE_i.
$$

The `boot::cv.glm()` function performs LOOCV when `K` is omitted. A
linear model can be fit with `glm()` using its default Gaussian family.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
loocv_error <- function(degree) {
  fit <- glm(mpg ~ poly(horsepower, degree), data = auto)
  cv.glm(auto, fit)$delta[1]
}

sapply(1:5, loocv_error)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.model_selection import LeaveOneOut, cross_val_score

def loocv_error(degree):
    model = make_pipeline(
        PolynomialFeatures(degree, include_bias=False),
        LinearRegression()
    )
    scores = cross_val_score(
        model,
        auto[["horsepower"]],
        auto["mpg"],
        cv=LeaveOneOut(),
        scoring="neg_mean_squared_error"
    )
    return -scores.mean()

print([loocv_error(degree) for degree in range(1, 6)])
```

:::

LOOCV uses nearly all observations for every fit, so it tends to have
low bias as an estimate of test error. Its disadvantage is computation:
large data sets or slow models can make LOOCV expensive.

## k-Fold Cross-Validation

k-fold cross-validation divides the data into $k$ roughly equal folds.
For each fold, fit the model on the other $k - 1$ folds and evaluate on
the held-out fold. The estimate is

$$
CV_{(k)} = \frac{1}{k}\sum_{j=1}^{k} MSE_j.
$$

LOOCV is the special case where $k=n$. In applied work, $k=5$ or
$k=10$ is often a good compromise between statistical stability and
computational cost.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
kfold_error <- function(degree, k = 10) {
  fit <- glm(mpg ~ poly(horsepower, degree), data = auto)
  cv.glm(auto, fit, K = k)$delta[1]
}

data.frame(
  degree = 1:5,
  loocv = sapply(1:5, loocv_error),
  ten_fold = sapply(1:5, kfold_error)
)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.model_selection import KFold

def kfold_error(degree, k=10):
    model = make_pipeline(
        PolynomialFeatures(degree, include_bias=False),
        LinearRegression()
    )
    scores = cross_val_score(
        model,
        auto[["horsepower"]],
        auto["mpg"],
        cv=KFold(n_splits=k, shuffle=True, random_state=571),
        scoring="neg_mean_squared_error"
    )
    return -scores.mean()

print(pd.DataFrame({
    "degree": range(1, 6),
    "loocv": [loocv_error(degree) for degree in range(1, 6)],
    "ten_fold": [kfold_error(degree) for degree in range(1, 6)]
}))
```

:::

The LOOCV and 10-fold estimates should tell the same broad story:
quadratic curvature matters, while higher-degree terms should be
justified by their estimated out-of-sample performance rather than by
training fit alone.

## Comparing Candidate Sales Models

The same logic applies to ordinary regression workflows. The `Carseats`
data contain simulated sales of child car seats at 400 stores. Suppose
we want to compare a full model with all predictors against a reduced
model that focuses on predictors with a clearer sales mechanism.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
carseats <- ISLR::Carseats

set.seed(571)
train_id <- sample(seq_len(nrow(carseats)), size = 0.8 * nrow(carseats))
train <- carseats[train_id, ]
test <- carseats[-train_id, ]

full_fit <- lm(Sales ~ ., data = train)
reduced_fit <- lm(
  Sales ~ CompPrice + Income + Advertising + Price + ShelveLoc + Age,
  data = train
)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
rng = np.random.default_rng(571)
carseats = pd.DataFrame({
    "Sales": rng.normal(8, 2, 400),
    "CompPrice": rng.normal(125, 15, 400),
    "Income": rng.normal(70, 25, 400),
    "Advertising": rng.uniform(0, 25, 400),
    "Price": rng.normal(115, 20, 400),
    "ShelveLoc": rng.choice(["Bad", "Medium", "Good"], 400),
    "Age": rng.normal(55, 15, 400),
    "Population": rng.normal(250, 100, 400)
})

train, test = train_test_split(carseats, train_size=0.8, random_state=571)
full_fit = smf.ols(
    "Sales ~ CompPrice + Income + Advertising + Price + C(ShelveLoc) + Age + Population",
    data=train
).fit()
reduced_fit = smf.ols(
    "Sales ~ CompPrice + Income + Advertising + Price + C(ShelveLoc) + Age",
    data=train
).fit()
```

:::

Residual standard error describes in-sample residual size. Test-set
RMSE and test-set $R^2$ describe predictive performance on held-out
observations.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
model_metrics <- function(fit, test_data, model_name) {
  pred <- predict(fit, newdata = test_data)
  actual <- test_data$Sales

  rmse <- sqrt(mean((actual - pred)^2))
  test_r2 <- 1 - sum((actual - pred)^2) /
    sum((actual - mean(actual))^2)

  data.frame(
    model = model_name,
    residual_se = sigma(fit),
    rmse = rmse,
    test_r2 = test_r2
  )
}

rbind(
  model_metrics(full_fit, test, "Full model"),
  model_metrics(reduced_fit, test, "Reduced model")
)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
def model_metrics(fit, test_data, model_name):
    pred = fit.predict(test_data)
    actual = test_data["Sales"]
    rmse = mean_squared_error(actual, pred) ** 0.5
    test_r2 = 1 - ((actual - pred) ** 2).sum() / ((actual - actual.mean()) ** 2).sum()
    return {
        "model": model_name,
        "residual_se": np.sqrt(fit.scale),
        "rmse": rmse,
        "test_r2": test_r2
    }

print(pd.DataFrame([
    model_metrics(full_fit, test, "Full model"),
    model_metrics(reduced_fit, test, "Reduced model")
]))
```

:::

Because the reduced model is nested inside the full model, an ANOVA
comparison can ask whether the omitted predictors improve the fit after
the reduced model's predictors are already included.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
anova(reduced_fit, full_fit)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from statsmodels.stats.anova import anova_lm

print(anova_lm(reduced_fit, full_fit))
```

:::

Model comparison should combine statistical evidence, test performance,
and interpretability. A slightly larger model is not automatically
better if its added variables do not improve generalization or clarify
the substantive claim.

Cross-validation lets us compare the same two models without relying on
one train/test split:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
glm_full <- glm(Sales ~ ., data = carseats)
glm_reduced <- glm(
  Sales ~ CompPrice + Income + Advertising + Price + ShelveLoc + Age,
  data = carseats
)

data.frame(
  model = c("Full model", "Reduced model"),
  cv_5 = c(
    cv.glm(carseats, glm_full, K = 5)$delta[1],
    cv.glm(carseats, glm_reduced, K = 5)$delta[1]
  ),
  cv_10 = c(
    cv.glm(carseats, glm_full, K = 10)$delta[1],
    cv.glm(carseats, glm_reduced, K = 10)$delta[1]
  ),
  loocv = c(
    cv.glm(carseats, glm_full)$delta[1],
    cv.glm(carseats, glm_reduced)$delta[1]
  )
)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.model_selection import cross_val_score
from sklearn.preprocessing import OneHotEncoder
from sklearn.compose import ColumnTransformer
from sklearn.pipeline import make_pipeline

X_full = carseats.drop(columns=["Sales"])
y_sales = carseats["Sales"]
X_reduced = carseats[["CompPrice", "Income", "Advertising", "Price", "ShelveLoc", "Age"]]

def cv_rmse(X, y, k):
    categorical = X.select_dtypes(exclude="number").columns
    preprocess = ColumnTransformer([
        ("categorical", OneHotEncoder(handle_unknown="ignore"), categorical)
    ], remainder="passthrough")
    model = make_pipeline(preprocess, LinearRegression())
    scores = cross_val_score(
        model, X, y,
        cv=KFold(n_splits=k, shuffle=True, random_state=571),
        scoring="neg_mean_squared_error"
    )
    return (-scores.mean()) ** 0.5

print(pd.DataFrame({
    "model": ["Full model", "Reduced model"],
    "cv_5": [cv_rmse(X_full, y_sales, 5), cv_rmse(X_reduced, y_sales, 5)],
    "cv_10": [cv_rmse(X_full, y_sales, 10), cv_rmse(X_reduced, y_sales, 10)]
}))
```

:::

## Simulated Nonlinear Data

A simulation makes the role of model complexity visible because we know
the data were generated from a nonlinear pattern.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
set.seed(571)
sim <- data.frame(x = runif(100, min = -10, max = 10))
sim$y <- sin(sim$x) + 0.1 * sim$x^2 + rnorm(100)

plot(y ~ x, data = sim, pch = 19, col = "#2f5f7e")
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
rng = np.random.default_rng(571)
sim = pd.DataFrame({"x": rng.uniform(-10, 10, 100)})
sim["y"] = np.sin(sim["x"]) + 0.1 * sim["x"] ** 2 + rng.normal(size=100)

plt.scatter(sim["x"], sim["y"], color="#2f5f7e")
plt.xlabel("x")
plt.ylabel("y")
plt.show()
```

:::

Fit polynomial models and use 10-fold cross-validation to estimate
generalization error:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
poly_cv <- sapply(1:15, function(degree) {
  fit <- glm(y ~ poly(x, degree), data = sim)
  cv.glm(sim, fit, K = 10)$delta[1]
})

plot(
  1:15,
  poly_cv,
  type = "b",
  xlab = "Polynomial degree",
  ylab = "10-fold CV error"
)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
poly_cv = [
    -cross_val_score(
        make_pipeline(PolynomialFeatures(degree, include_bias=False), LinearRegression()),
        sim[["x"]],
        sim["y"],
        cv=KFold(n_splits=10, shuffle=True, random_state=571),
        scoring="neg_mean_squared_error"
    ).mean()
    for degree in range(1, 16)
]

plt.plot(range(1, 16), poly_cv, marker="o")
plt.xlabel("Polynomial degree")
plt.ylabel("10-fold CV error")
plt.show()
```

:::

This exercise separates visual fit from estimated generalization. A
model with many degrees may trace the observed points closely, but
cross-validation asks whether that flexibility is likely to help on new
data.

## Bootstrapping

Cross-validation estimates prediction error. Bootstrapping estimates
uncertainty. A bootstrap sample draws $n$ rows from the original data,
with replacement, and recomputes a statistic on the sampled data. After
many bootstrap samples, the spread of the bootstrap estimates describes
the statistic's sampling variability.

If $\hat\alpha^{*1}, \ldots, \hat\alpha^{*B}$ are bootstrap estimates,
the bootstrap standard error is

$$
SE_B(\hat\alpha)
= \sqrt{
\frac{1}{B-1}
\sum_{r=1}^{B}
\left(
\hat\alpha^{*r}
- \frac{1}{B}\sum_{s=1}^{B}\hat\alpha^{*s}
\right)^2
}.
$$

For the `Portfolio` data, one statistic of interest is the allocation
that minimizes variance across two assets:

$$
\hat\alpha =
\frac{\hat\sigma_Y^2 - \hat\sigma_{XY}}
{\hat\sigma_X^2 + \hat\sigma_Y^2 - 2\hat\sigma_{XY}}.
$$

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
portfolio <- ISLR::Portfolio

alpha_stat <- function(data, index) {
  x <- data$X[index]
  y <- data$Y[index]
  (var(y) - cov(x, y)) / (var(x) + var(y) - 2 * cov(x, y))
}

set.seed(571)
boot(portfolio, alpha_stat, R = 1000)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
rng = np.random.default_rng(571)
portfolio = pd.DataFrame({
    "X": rng.normal(0, 1, 100),
    "Y": rng.normal(0, 1, 100)
})

def alpha_stat(data):
    x = data["X"]
    y = data["Y"]
    return (y.var() - x.cov(y)) / (x.var() + y.var() - 2 * x.cov(y))

boot_alpha = []
for _ in range(1000):
    sample = portfolio.sample(len(portfolio), replace=True, random_state=rng.integers(1_000_000))
    boot_alpha.append(alpha_stat(sample))

print(pd.Series(boot_alpha).describe())
```

:::

The same idea applies to regression coefficients. The function below
refits the model on each bootstrap sample and returns the intercept and
slope.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
boot_lm <- function(data, index) {
  coef(lm(mpg ~ horsepower, data = data, subset = index))
}

set.seed(571)
boot(auto, boot_lm, R = 1000)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
boot_lm = []
for _ in range(1000):
    sample = auto.sample(len(auto), replace=True, random_state=rng.integers(1_000_000))
    fit = smf.ols("mpg ~ horsepower", data=sample).fit()
    boot_lm.append(fit.params)

print(pd.DataFrame(boot_lm).describe())
```

:::

If a nonlinear mean structure is more appropriate, bootstrap the
quadratic model instead:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
boot_quad <- function(data, index) {
  coef(lm(mpg ~ poly(horsepower, 2), data = data, subset = index))
}

set.seed(571)
boot(auto, boot_quad, R = 1000)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
boot_quad = []
for _ in range(1000):
    sample = auto.sample(len(auto), replace=True, random_state=rng.integers(1_000_000))
    sample = sample.assign(horsepower2=sample["horsepower"] ** 2)
    fit = smf.ols("mpg ~ horsepower + horsepower2", data=sample).fit()
    boot_quad.append(fit.params)

print(pd.DataFrame(boot_quad).describe())
```

:::

The bootstrap does not replace model criticism. It quantifies
variability for the statistic implied by the model and the data
collection process. If the model is misspecified or the data are not
representative, the bootstrap can still give a precise answer to the
wrong question.
