---
title: "Data Splitting and Feature Engineering"
---

Statistical learning depends on disciplined data handling. A model can
look strong simply because information from the test data leaked into
preprocessing, feature construction, or tuning. This chapter focuses on
building model-ready data without contaminating the evidence used for
evaluation.

## Learning Objectives

- Create training, validation, and test partitions.
- Explain why preprocessing must be estimated inside the training data.
- Encode categorical variables and transform skewed predictors.
- Construct interactions and nonlinear features deliberately.
- Recognize leakage in common modeling workflows.

## Splitting Comes First

A split creates a contract. The training data are used to estimate
preprocessing steps, fit models, and tune candidate choices. The test
data are held back for final evaluation.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(rsample)
library(AmesHousing)
library(modeldata)

# To get the processed version:
ames <- make_ames()

set.seed(571)
ames_split <- initial_split(ames, prop = 0.8, strata = Sale_Price)
ames_train <- training(ames_split)
ames_test <- testing(ames_split)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.model_selection import train_test_split

ames_train, ames_test = train_test_split(
    ames,
    train_size=0.8,
    random_state=571
)
```

:::

Stratification keeps the response distribution roughly comparable
across partitions. This is useful when the response is skewed or when a
class outcome is imbalanced.

## Leakage

Leakage occurs when information unavailable at prediction time enters
the model-building process. Common examples include:

| Step | Leakage risk |
|---|---|
| Scaling before splitting | Test-set means and standard deviations influence training data |
| Imputing before splitting | Test-set missingness affects imputation values |
| Selecting predictors before splitting | Test-set associations influence the model space |
| Tuning on the test set | Final evaluation becomes part of model selection |
| Using future variables | Predictors contain information not known at decision time |

The anti-leakage rule is simple: learn preprocessing from the training
data, then apply the learned transformation to validation, test, and
future data.

## Ames Housing Workflow

For Ames housing, `Sale_Price` is the target and many predictors are
categorical, skewed, sparse, or missing. A reproducible feature
workflow should handle all of these in a fixed order.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(recipes)

ames_recipe <-
  recipe(Sale_Price ~ ., data = ames_train) |>
  step_log(Sale_Price, base = 10, skip = TRUE) |>
  step_other(all_nominal_predictors(), threshold = 0.01) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

ames_prep <- prep(ames_recipe, training = ames_train)
train_ready <- bake(ames_prep, new_data = NULL)
test_ready <- bake(ames_prep, new_data = ames_test)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
from sklearn.compose import ColumnTransformer
from sklearn.impute import SimpleImputer
from sklearn.pipeline import Pipeline
from sklearn.preprocessing import OneHotEncoder, StandardScaler

y_train = np.log10(ames_train["Sale_Price"])
X_train = ames_train.drop(columns=["Sale_Price"])
X_test = ames_test.drop(columns=["Sale_Price"])

numeric_features = X_train.select_dtypes(include="number").columns
categorical_features = X_train.select_dtypes(exclude="number").columns

numeric_pipeline = Pipeline([
    ("impute", SimpleImputer(strategy="median")),
    ("scale", StandardScaler())
])
categorical_pipeline = Pipeline([
    ("impute", SimpleImputer(strategy="most_frequent")),
    ("dummy", OneHotEncoder(handle_unknown="ignore"))
])

preprocess = ColumnTransformer([
    ("numeric", numeric_pipeline, numeric_features),
    ("categorical", categorical_pipeline, categorical_features)
])

train_ready = preprocess.fit_transform(X_train)
test_ready = preprocess.transform(X_test)
```

:::

The recipe is estimated on `ames_train`. The test data receive the same
learned mapping: the same rare-level rules, dummy columns, zero-variance
filter, and normalization constants.

## Transformations

Transformations change the representation of predictors or the
response. They can make simple models more effective, but they also
change interpretation.

| **Transformation** | **When useful **| **Interpretation** |
|---|---|---|
| Log response | Positive, right-skewed outcome | Effects are multiplicative on the original scale |
| Log predictor | Diminishing returns in a positive predictor | One percent changes become natural |
| Standardization | Penalties, distance methods, PCA, SVMs | Coefficients use standard deviation units |
| Binning | Highly nonlinear or threshold-like behavior | Interpretation becomes interval-based |
| Dummy encoding | Nominal categorical predictors | Coefficients compare to a reference level |

Transformations should be chosen for a modeling reason, not merely
because they make a plot look smoother.

## Interaction Features

An interaction is feature construction with a statistical
interpretation. In the model

$$
\operatorname{E}(Y \mid X,Z)
= \beta_0 + \beta_1X + \beta_2Z + \beta_3XZ,
$$

the effect of $X$ changes with $Z$. Interactions are especially useful
when theory or visualization suggests that a relationship differs by
group or by level of another variable.

In housing data, the effect of living area might differ by neighborhood
or overall quality. In advertising data, the return to TV spending
might depend on radio spending. Interactions should be evaluated with
held-out error and interpreted with marginal effects or predicted
values, not only with coefficient signs.

## Model-Ready Data Is Not Model-Free

Preprocessing embeds modeling choices. Scaling assumes that relative
distance matters. Rare-level lumping assumes sparse categories should
share information. Dummy encoding chooses a reference structure.
Feature engineering is therefore part of the model, and it must be
reported with the same care as the final algorithm.
