---
title: "Logistic Regression and ROC Analysis"
---

Logistic regression is the core statistical learning model for binary
outcomes. It estimates probabilities, supports interpretable
coefficients, and introduces the evaluation tools used throughout
classification.

## Learning Objectives

- Fit a logistic regression model.
- Interpret log odds, odds ratios, and fitted probabilities.
- Use bins to inspect monotone risk patterns.
- Compare thresholds with sensitivity and specificity.
- Use ROC curves and AUC to evaluate ranking performance.

## The Logistic Model

For a binary outcome,

$$
p(x) = \Pr(Y=1 \mid X=x).
$$

Logistic regression models the log odds:

$$
\log\left(\frac{p(x)}{1-p(x)}\right)
= \beta_0 + x^\top\beta.
$$

The inverse-logit transformation returns the probability:

$$
p(x) =
\frac{\exp(\beta_0+x^\top\beta)}
{1+\exp(\beta_0+x^\top\beta)}.
$$

## Credit Risk Example

Credit-risk modeling makes three ideas visible: rare events,
probability estimation, and threshold choice.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
set.seed(571)
n <- 1000
credit <- data.frame(
  balance = rgamma(n, shape = 4, scale = 350),
  income = rnorm(n, mean = 60000, sd = 18000),
  age = round(rnorm(n, mean = 45, sd = 12))
)
eta <- -6 + 0.0032 * credit$balance -
  0.000012 * credit$income - 0.015 * credit$age
credit$default <- rbinom(n, 1, plogis(eta))

logit_fit <- glm(default ~ balance + income + age,
                 data = credit,
                 family = binomial)
credit$prob_default <- predict(logit_fit, type = "response")
summary(logit_fit)
exp(coef(logit_fit))
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
import pandas as pd
import statsmodels.formula.api as smf

rng = np.random.default_rng(571)
n = 1000
credit = pd.DataFrame({
    "balance": rng.gamma(shape=4, scale=350, size=n),
    "income": rng.normal(60000, 18000, n),
    "age": np.round(rng.normal(45, 12, n))
})
eta = -6 + 0.0032 * credit["balance"] - 0.000012 * credit["income"] - 0.015 * credit["age"]
credit["default"] = rng.binomial(1, 1 / (1 + np.exp(-eta)))

logit_fit = smf.logit("default ~ balance + income + age", data=credit).fit()
credit["prob_default"] = logit_fit.predict(credit)
print(logit_fit.summary())
print(np.exp(logit_fit.params))
```

:::

An odds ratio above 1 increases odds of default; an odds ratio below 1
decreases odds, holding other predictors fixed. Probability changes
are not constant because the logit curve is nonlinear.

## Binning and Default Rates

Binning a continuous predictor can reveal whether observed risk changes
monotonically with the predictor.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
credit$balance_bin <- cut(credit$balance,
                          breaks = quantile(credit$balance, probs = seq(0, 1, 0.1)),
                          include.lowest = TRUE)
aggregate(default ~ balance_bin, data = credit, mean)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
credit["balance_bin"] = pd.qcut(credit["balance"], q=10, duplicates="drop")
print(credit.groupby("balance_bin")["default"].mean())
```

:::

Binning is not a replacement for modeling, but it is a useful
diagnostic for risk monotonicity and communication.

## ROC and AUC

An ROC curve plots sensitivity against the false positive rate across
all possible thresholds. AUC summarizes how well the model ranks
positive cases above negative cases.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
thresholds <- seq(0, 1, length.out = 101)
roc_points <- do.call(rbind, lapply(thresholds, function(cutoff) {
  pred <- ifelse(credit$prob_default >= cutoff, 1, 0)
  tp <- sum(pred == 1 & credit$default == 1)
  fp <- sum(pred == 1 & credit$default == 0)
  tn <- sum(pred == 0 & credit$default == 0)
  fn <- sum(pred == 0 & credit$default == 1)
  data.frame(
    threshold = cutoff,
    sensitivity = tp / (tp + fn),
    false_positive_rate = fp / (fp + tn)
  )
}))
head(roc_points)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
thresholds = np.linspace(0, 1, 101)
records = []
for cutoff in thresholds:
    pred = (credit["prob_default"] >= cutoff).astype(int)
    tp = ((pred == 1) & (credit["default"] == 1)).sum()
    fp = ((pred == 1) & (credit["default"] == 0)).sum()
    tn = ((pred == 0) & (credit["default"] == 0)).sum()
    fn = ((pred == 0) & (credit["default"] == 1)).sum()
    records.append({
        "threshold": cutoff,
        "sensitivity": tp / (tp + fn),
        "false_positive_rate": fp / (fp + tn)
    })
roc_points = pd.DataFrame(records)
print(roc_points.head())
```

:::

ROC curves describe ranking performance, not calibration. A model can
rank cases well while producing probabilities that are too high or too
low.
