---
title: "Model Selection and Regularization"
---

Model selection asks which predictors and which level of complexity
should be used. Regularization answers the same question by changing
the estimation problem: instead of choosing one subset of predictors
directly, it penalizes coefficient size or coefficient count.

## Learning Objectives

- Compare best subset, forward stepwise, and backward stepwise
  selection.
- Use adjusted $R^2$, $C_p$, BIC, AIC, validation error, and
  cross-validation error.
- Explain ridge regression and lasso as penalized least squares.
- Describe why standardization matters for penalized regression.
- Use cross-validation to choose a tuning parameter.

## Why Selection Is Hard

With $p$ predictors, best subset selection considers $2^p$ possible
models. That search can find a model that fits the training data well,
but the search itself adds variability. The selected model may look
more certain than it really is because the same data were used to try
many alternatives.

Selection should therefore be evaluated with a criterion that accounts
for complexity or with held-out prediction error.

## Subset and Stepwise Selection

Best subset selection chooses the best model of each size. Forward
stepwise selection starts from the intercept-only model and adds one
predictor at a time. Backward stepwise selection starts from the full
model and removes one predictor at a time.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(leaps)

subset_fit <- regsubsets(Sale_Price ~ ., data = train, nvmax = 20)
subset_summary <- summary(subset_fit)

which.min(subset_summary$bic)
which.max(subset_summary$adjr2)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import itertools
import numpy as np
import pandas as pd
import statsmodels.api as sm

predictors = [col for col in train.columns if col != "Sale_Price"]
records = []

for size in range(1, min(len(predictors), 5) + 1):
    for cols in itertools.combinations(predictors, size):
        X = sm.add_constant(pd.get_dummies(train[list(cols)], drop_first=True))
        fit = sm.OLS(train["Sale_Price"], X).fit()
        records.append({
            "size": size,
            "predictors": cols,
            "aic": fit.aic,
            "bic": fit.bic,
            "adj_r2": fit.rsquared_adj
        })

subset_summary = pd.DataFrame(records)
print(subset_summary.sort_values("bic").head())
print(subset_summary.sort_values("adj_r2", ascending=False).head())
```

:::

The resulting model path can be compared using:

| Criterion | Preference | Caution |
|---|---|---|
| Training RSS | Larger models | Always decreases as predictors are added |
| Adjusted $R^2$ | Fit adjusted for model size | Still an in-sample measure |
| Mallows' $C_p$ | Low estimated prediction error | Depends on variance estimate |
| AIC | Predictive fit with lighter penalty | Can select larger models |
| BIC | Fit with stronger penalty | Often selects simpler models |
| Cross-validation | Held-out predictive performance | Depends on split design |

## Media Incivility Selection Example

A model-selection analysis should connect the criterion to the research
question. For a media-incivility outcome with many political,
platform, or text features, an exhaustive search can identify candidate
models of each size, while forward and backward selection provide
computationally cheaper alternatives.

The analysis workflow is:

1. define the outcome and candidate predictors,
2. split the data or define cross-validation folds,
3. fit the candidate model paths on the training data,
4. compare $R^2$, adjusted $R^2$, $C_p$, BIC, and validation error,
5. refit the selected specification on the training data,
6. evaluate once on the held-out data,
7. describe the selected predictors without overstating certainty.

## Ridge Regression

Ridge regression minimizes

$$
\sum_{i=1}^{n}(y_i - x_i^\top\beta)^2
+ \lambda \sum_{j=1}^{p}\beta_j^2.
$$

The $\ell_2$ penalty shrinks coefficients toward zero but usually does
not set them exactly to zero. Ridge is especially helpful when
predictors are correlated: it distributes signal across related
variables and lowers variance.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(glmnet)

x_train <- model.matrix(Apps ~ ., College)[, -1]
y_train <- College$Apps

ridge_cv <- cv.glmnet(x_train, y_train, alpha = 0)
coef(ridge_cv, s = "lambda.min")
coef(ridge_cv, s = "lambda.1se")
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
import pandas as pd
from sklearn.linear_model import RidgeCV
from sklearn.preprocessing import StandardScaler
from sklearn.pipeline import make_pipeline

x_train = pd.get_dummies(college.drop(columns=["Apps"]), drop_first=True)
y_train = college["Apps"]

ridge_cv = make_pipeline(
    StandardScaler(),
    RidgeCV(alphas=np.logspace(-3, 3, 100))
)
ridge_cv.fit(x_train, y_train)

print(ridge_cv.named_steps["ridgecv"].alpha_)
```

:::

## Lasso

The lasso minimizes

$$
\sum_{i=1}^{n}(y_i - x_i^\top\beta)^2
+ \lambda \sum_{j=1}^{p}|\beta_j|.
$$

The $\ell_1$ penalty can set coefficients exactly to zero, so lasso is
both a prediction method and a variable-selection method. When
predictors are strongly correlated, lasso may choose one representative
predictor and ignore others with similar information.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
lasso_cv <- cv.glmnet(x_train, y_train, alpha = 1)
coef(lasso_cv, s = "lambda.min")
coef(lasso_cv, s = "lambda.1se")
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.linear_model import LassoCV
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import StandardScaler

lasso_cv = make_pipeline(
    StandardScaler(),
    LassoCV(cv=10, random_state=571, max_iter=10000)
)
lasso_cv.fit(x_train, y_train)

print(lasso_cv.named_steps["lassocv"].alpha_)
```

:::

`lambda.min` gives the value with the lowest cross-validation error.
`lambda.1se` gives the simplest model whose error is within one
standard error of the minimum. The one-standard-error rule is often a
good teaching device because it makes the accuracy-simplicity tradeoff
explicit.

## Scaling and Interpretation

Penalties operate on coefficient size, so predictors must be on
comparable scales. Otherwise a variable measured in large units may be
penalized differently from a variable measured in small units. Most
software standardizes predictors internally by default, but analysts
should know whether and how that standardization occurred.

After penalized regression, coefficient interpretation should be
careful. Shrinkage intentionally biases coefficients toward zero to
reduce variance. Penalized coefficients are not ordinary least-squares
estimates with ordinary standard errors.
