---
title: "Dimension Reduction"
---

Dimension reduction replaces many correlated variables with a smaller
set of constructed features. The goal may be visualization,
compression, noise reduction, or prediction when the original predictor
space is too large or too collinear for stable ordinary regression.

## Learning Objectives

- Explain principal components as directions of maximal variance.
- Distinguish PCA, principal components regression, and partial least
  squares.
- Explain why centering and scaling change principal components.
- Use cross-validation to choose the number of components.
- Interpret component loadings and prediction results together.

## Principal Components

Let $X$ be a centered and scaled predictor matrix. Principal component
analysis finds loading vectors $v_1, v_2,\ldots$ so that

$$
z_1 = Xv_1
$$

has the largest possible variance subject to $\|v_1\|=1$. The second
component maximizes variance subject to being orthogonal to the first,
and so on. The score vectors $z_1,z_2,\ldots$ are new predictors:
uncorrelated linear combinations of the original variables.

PCA is unsupervised. It uses only $X$, not the response $Y$.

## Scaling Is a Modeling Choice

PCA is sensitive to measurement units. If one predictor has values in
thousands and another has values in fractions, the first can dominate
the components even if it is not more important. Centering removes
location differences; scaling puts variables on comparable units.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
college <- ISLR::College
x <- model.matrix(Apps ~ ., college)[, -1]

pca_unscaled <- prcomp(x, center = TRUE, scale. = FALSE)
pca_scaled <- prcomp(x, center = TRUE, scale. = TRUE)

summary(pca_scaled)
biplot(pca_scaled, scale = 0)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import pandas as pd
from sklearn.decomposition import PCA
from sklearn.preprocessing import StandardScaler

x = pd.get_dummies(college.drop(columns=["Apps"]), drop_first=True)

pca_unscaled = PCA().fit(x)
x_scaled = StandardScaler().fit_transform(x)
pca_scaled = PCA().fit(x_scaled)

print(pca_scaled.explained_variance_ratio_[:5])
```

:::

An incorrect workflow scales the full data set before splitting, which
leaks information. A correct workflow estimates centering, scaling, and
components from the training data, then applies the learned rotation to
held-out data.

## Principal Components Regression

Principal components regression first constructs components from the
predictors, then regresses the outcome on the first $M$ components.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(pls)

set.seed(571)
pcr_fit <- pcr(
  Apps ~ .,
  data = college,
  scale = TRUE,
  validation = "CV"
)

validationplot(pcr_fit, val.type = "MSEP")
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
from sklearn.decomposition import PCA
from sklearn.linear_model import LinearRegression
from sklearn.model_selection import cross_val_score
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import StandardScaler

component_grid = range(1, 11)
pcr_scores = []

for n_components in component_grid:
    pcr = make_pipeline(
        StandardScaler(),
        PCA(n_components=n_components),
        LinearRegression()
    )
    mse = -cross_val_score(
        pcr, x, college["Apps"],
        scoring="neg_mean_squared_error",
        cv=10
    ).mean()
    pcr_scores.append(mse)

print(list(zip(component_grid, pcr_scores)))
```

:::

Keeping too few components can miss predictive signal. Keeping too many
components can reintroduce noise. Cross-validation provides an empirical
way to choose $M$.

## Partial Least Squares

Partial least squares also constructs components, but it uses the
response while constructing them. This can help when directions with
large predictor variance are not the directions most related to the
outcome.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
pls_fit <- plsr(
  Apps ~ .,
  data = college,
  scale = TRUE,
  validation = "CV"
)

validationplot(pls_fit, val.type = "MSEP")
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.cross_decomposition import PLSRegression
from sklearn.model_selection import cross_val_score
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import StandardScaler

pls_scores = []

for n_components in component_grid:
    pls = make_pipeline(
        StandardScaler(),
        PLSRegression(n_components=n_components)
    )
    mse = -cross_val_score(
        pls, x, college["Apps"],
        scoring="neg_mean_squared_error",
        cv=10
    ).mean()
    pls_scores.append(mse)

print(list(zip(component_grid, pls_scores)))
```

:::

| Method | Uses $Y$ to construct components? | Main purpose |
|---|---:|---|
| PCA | No | Exploration and compression |
| PCR | No, then regression | Prediction with correlated predictors |
| PLS | Yes | Prediction when response-related directions matter |

## Near-Infrared Spectra Example

High-dimensional spectra data illustrate why dimension reduction is
useful. Each observation can contain hundreds of wavelength
measurements, with neighboring wavelengths strongly correlated. A small
number of components can summarize broad spectral patterns while
reducing noise.

The analysis sequence is:

1. center and scale spectral predictors inside the training data,
2. visualize the first two component scores,
3. inspect loadings to see which wavelengths define each component,
4. fit PCR and PLS models,
5. choose the number of components by cross-validation,
6. evaluate the selected model on held-out observations.

This example also shows that interpretability changes. Instead of
interpreting one coefficient per wavelength, students interpret
component loadings and prediction performance together.
