---
title: "Bias, Variance, and Nonlinear Models"
---

This chapter connects model flexibility to prediction error and then
uses nonlinear regression tools to show how simple models can learn
curved relationships without abandoning statistical interpretation.

## Learning Objectives

- Explain reducible error, irreducible error, bias, and variance.
- Identify underfitting and overfitting from train/test patterns.
- Fit polynomial regression, step functions, splines, local regression,
  and generalized additive models.
- Use validation or cross-validation to choose flexibility.
- Explain nonlinear fitted relationships with plots and predicted
  values.

## Bias-Variance Decomposition

At a new point $x_0$, expected prediction error can be decomposed as

$$
\operatorname{E}\{(Y - \hat f(x_0))^2\}
= \operatorname{Var}(\epsilon)
+ \operatorname{Bias}(\hat f(x_0))^2
+ \operatorname{Var}(\hat f(x_0)).
$$

The irreducible error $\operatorname{Var}(\epsilon)$ remains even if
the true regression function is known. Bias is error from using a model
that is too rigid. Variance is error from using a model that changes
too much across training samples.

Increasing flexibility usually lowers bias and raises variance. The
best predictive model is the one that balances these forces for the
available sample size and signal strength.

## Simulation View

A simulation makes the tradeoff visible because the true function can
be known:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
set.seed(571)
x <- runif(200, -3, 3)
y <- sin(x) + 0.4 * x^2 + rnorm(200, sd = 0.4)

fit_linear <- lm(y ~ x)
fit_poly5 <- lm(y ~ poly(x, 5))
fit_poly15 <- lm(y ~ poly(x, 15))
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
import pandas as pd
from sklearn.linear_model import LinearRegression
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import PolynomialFeatures

rng = np.random.default_rng(571)
x = rng.uniform(-3, 3, 200)
y = np.sin(x) + 0.4 * x**2 + rng.normal(0, 0.4, 200)
X = x.reshape(-1, 1)

fit_linear = LinearRegression().fit(X, y)
fit_poly5 = make_pipeline(PolynomialFeatures(5), LinearRegression()).fit(X, y)
fit_poly15 = make_pipeline(PolynomialFeatures(15), LinearRegression()).fit(X, y)
```

:::

The linear model has high bias if the true function is curved. The
degree-15 polynomial may have high variance if it follows accidental
noise. Validation error, not visual smoothness alone, should guide the
choice.

## Polynomial Regression

Polynomial regression remains linear in the coefficients while allowing
a curved mean function:

$$
\operatorname{E}(Y \mid X=x)
= \beta_0 + \beta_1x + \beta_2x^2 + \cdots + \beta_dx^d.
$$

In R, orthogonal polynomial terms reduce numerical collinearity:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(ISLR)

wage <- ISLR::Wage
poly_fit <- lm(wage ~ poly(age, 4), data = wage)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.linear_model import LinearRegression
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import PolynomialFeatures

X_age = wage[["age"]]
y_wage = wage["wage"]

poly_fit = make_pipeline(
    PolynomialFeatures(degree=4, include_bias=False),
    LinearRegression()
).fit(X_age, y_wage)
```

:::

The degree controls flexibility. Prediction plots are usually clearer
than raw coefficient interpretation for high-degree polynomials.

## Step Functions and Splines

A step function divides a predictor into intervals and fits a constant
mean within each interval:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
step_fit <- lm(wage ~ cut(age, 4), data = wage)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import pandas as pd
import statsmodels.formula.api as smf

wage_step = wage.copy()
wage_step["age_bin"] = pd.cut(wage_step["age"], bins=4)
step_fit = smf.ols("wage ~ C(age_bin)", data=wage_step).fit()
print(step_fit.summary())
```

:::

Regression splines fit smooth piecewise polynomials. Knots define where
the pieces meet, and degrees of freedom control smoothness:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(splines)

spline_fit <- lm(wage ~ bs(age, df = 6), data = wage)
natural_fit <- lm(wage ~ ns(age, df = 5), data = wage)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import statsmodels.formula.api as smf

spline_fit = smf.ols(
    "wage ~ bs(age, df=6)",
    data=wage
).fit()
natural_fit = smf.ols(
    "wage ~ cr(age, df=5)",
    data=wage
).fit()
```

:::

Natural splines add boundary constraints that often produce more stable
behavior at the edges of the data.

## Generalized Additive Models

Generalized additive models combine several smooth functions:

$$
g\{\operatorname{E}(Y \mid X)\}
= \beta_0 + f_1(X_1) + f_2(X_2) + \cdots + f_p(X_p).
$$

They keep additivity while allowing nonlinear effects:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(gam)

gam_fit <- gam(
  wage ~ s(age, df = 5) + s(year, df = 4) + education,
  data = wage
)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import statsmodels.formula.api as smf

gam_like_fit = smf.ols(
    "wage ~ bs(age, df=5) + bs(year, df=4) + C(education)",
    data=wage
).fit()
```

:::

GAM plots show the estimated contribution of each smooth term while
holding other terms in the model fixed.

## Choosing Flexibility

Training error usually decreases as flexibility increases. Test error
often decreases at first, then increases when the model starts fitting
noise. A reliable comparison should use a validation set or
cross-validation:

1. choose a grid of degrees, knots, or smoothness values,
2. fit each candidate using the training folds,
3. compute held-out error,
4. choose a flexible model only if it earns better generalization,
5. explain the fitted shape with prediction plots.
