---
title: "Classification Model Comparison"
---

Classification model comparison asks which probability model is most
useful for the decision at hand. Logistic regression, LDA, and QDA can
all return class probabilities, but they make different assumptions and
can behave differently under imbalance, nonlinear boundaries, and
threshold changes.

## Learning Objectives

- Compare logistic regression, LDA, and QDA on the same task.
- Distinguish probability quality from hard-label accuracy.
- Explain how thresholds change sensitivity and specificity.
- Use ROC curves, AUC, and calibration to compare classifiers.
- Choose a classifier based on the decision context rather than one
  metric alone.

## Logistic Regression Baseline

For a binary outcome, logistic regression models

$$
p(x) = \Pr(Y = 1 \mid X=x).
$$

The logit link maps probabilities to the real line:

$$
\log\left(\frac{p(x)}{1-p(x)}\right)
= \beta_0 + x^\top\beta.
$$

Equivalently,

$$
p(x) =
\frac{\exp(\beta_0 + x^\top\beta)}
{1 + \exp(\beta_0 + x^\top\beta)}.
$$

Coefficients are changes in log odds. Exponentiating a coefficient
gives an odds ratio, holding the other predictors fixed.

## Credit Default Example

The default data show why probability estimation matters. The response
is rare, so accuracy can be misleading: a classifier that predicts "no
default" for everyone may have high accuracy but no practical value.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(ISLR)

default_fit <- glm(
  default ~ student + balance + income,
  data = Default,
  family = binomial
)

summary(default_fit)
exp(coef(default_fit))
Default$prob_default <- predict(default_fit, type = "response")
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
import pandas as pd
import statsmodels.formula.api as smf

default = default.copy()
default["default_yes"] = (default["default"] == "Yes").astype(int)

default_fit = smf.logit(
    "default_yes ~ C(student) + balance + income",
    data=default
).fit()

print(default_fit.summary())
print(np.exp(default_fit.params))
default["prob_default"] = default_fit.predict(default)
```

:::

The fitted probabilities can be sorted, binned, and compared with
observed default rates. This helps students see whether high predicted
risk actually corresponds to higher observed risk.

## Thresholds and Error Costs

Logistic regression returns probabilities. A threshold turns
probabilities into class labels:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
pred_class <- ifelse(Default$prob_default > 0.5, "Yes", "No")
table(predicted = pred_class, observed = Default$default)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
pred_class = np.where(default["prob_default"] > 0.5, "Yes", "No")
print(pd.crosstab(pred_class, default["default"]))
```

:::

The threshold should reflect the decision context. If false negatives
are expensive, a lower threshold may be appropriate. If false positives
are expensive, a higher threshold may be appropriate. The statistical
model estimates risk; the threshold encodes a decision rule.

## LDA and QDA Baselines

Discriminant analysis starts by modeling the distribution of predictors
within each class. Bayes' rule gives

$$
\Pr(Y=k \mid X=x)
=
\frac{\pi_k f_k(x)}
{\sum_{\ell=1}^{K} \pi_\ell f_\ell(x)},
$$

where $\pi_k$ is the prior probability of class $k$ and $f_k(x)$ is the
class-conditional density.

LDA assumes that each class has a multivariate normal predictor
distribution with a common covariance matrix. This produces linear
decision boundaries. QDA allows a different covariance matrix in each
class, producing quadratic boundaries.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(MASS)

lda_fit <- lda(default ~ balance + income + student, data = Default)
qda_fit <- qda(default ~ balance + income + student, data = Default)

lda_pred <- predict(lda_fit)$posterior[, "Yes"]
qda_pred <- predict(qda_fit)$posterior[, "Yes"]
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.discriminant_analysis import LinearDiscriminantAnalysis
from sklearn.discriminant_analysis import QuadraticDiscriminantAnalysis
from sklearn.preprocessing import OneHotEncoder
from sklearn.compose import ColumnTransformer
from sklearn.pipeline import make_pipeline

X = default[["balance", "income", "student"]]
y = default["default"]

preprocess = ColumnTransformer([
    ("student", OneHotEncoder(drop="first"), ["student"])
], remainder="passthrough")

lda_fit = make_pipeline(preprocess, LinearDiscriminantAnalysis()).fit(X, y)
qda_fit = make_pipeline(preprocess, QuadraticDiscriminantAnalysis()).fit(X, y)

lda_pred = lda_fit.predict_proba(X)[:, list(lda_fit.classes_).index("Yes")]
qda_pred = qda_fit.predict_proba(X)[:, list(qda_fit.classes_).index("Yes")]
```

:::

LDA can be more stable when data are limited because it estimates fewer
parameters. QDA can be more flexible when class covariance patterns
really differ and there is enough data to estimate them.

## ROC, AUC, and Calibration

Classification evaluation should include more than one number:

| Measure | What it answers |
|---|---|
| Accuracy | What fraction of labels are correct at one threshold? |
| Sensitivity | What fraction of positives are detected? |
| Specificity | What fraction of negatives are correctly rejected? |
| ROC curve | How do sensitivity and specificity trade off across thresholds? |
| AUC | How well are positives ranked above negatives? |
| Calibration | Do predicted probabilities match observed frequencies? |

ROC curves evaluate ranking across thresholds. Calibration evaluates
probability quality. A model can rank well and still produce poorly
calibrated probabilities.
