---
title: "Tree-Based Methods"
---

Tree-based methods partition the predictor space into regions and make
simple predictions inside each region. A single tree is easy to explain
and capable of representing interactions and nonlinearities, but it can
be unstable. That instability motivates ensemble methods in the next
chapter.

## Learning Objectives

- Explain recursive binary splitting.
- Distinguish regression trees and classification trees.
- Interpret impurity, tree depth, terminal nodes, and pruning.
- Use cross-validation to choose tree complexity.
- Compare a tree with a linear or logistic baseline.

## Recursive Binary Splitting

A decision tree repeatedly splits the predictor space. At each step, it
chooses a variable and cut point that best improve the fit.

For regression trees, a common objective is reduction in residual sum
of squares. For classification trees, common impurity measures include
the Gini index

$$
G = \sum_{k=1}^{K}\hat p_{mk}(1-\hat p_{mk})
$$

and cross-entropy

$$
D = -\sum_{k=1}^{K}\hat p_{mk}\log(\hat p_{mk}),
$$

where $\hat p_{mk}$ is the proportion of class $k$ observations in node
$m$.

## Tree Structure

A fitted tree contains:

| Term | Meaning |
|---|---|
| Root node | The full training data before any split |
| Internal node | A region that is split again |
| Terminal node | A final region used for prediction |
| Depth | Number of split levels from root to leaf |
| Split rule | A predictor threshold or category partition |
| Prediction | Mean response or class distribution inside a terminal node |

Prediction follows a path of questions. This makes trees easy to
communicate to nontechnical audiences.

## Ames and Carseats Examples

In a housing example, a regression tree may split first on overall
quality or living area, then on neighborhood, age, or basement size. In
a car-seat sales example, a tree may split on shelving location, price,
advertising, or competitor price.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(rpart)

sales_tree <- rpart(Sales ~ ., data = ISLR::Carseats)
printcp(sales_tree)
plot(sales_tree)
text(sales_tree, use.n = TRUE)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import pandas as pd
from sklearn.compose import ColumnTransformer
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import OneHotEncoder
from sklearn.tree import DecisionTreeRegressor, plot_tree
import matplotlib.pyplot as plt

X = carseats.drop(columns=["Sales"])
y = carseats["Sales"]
categorical = X.select_dtypes(exclude="number").columns

preprocess = ColumnTransformer([
    ("categorical", OneHotEncoder(handle_unknown="ignore"), categorical)
], remainder="passthrough")

sales_tree = make_pipeline(
    preprocess,
    DecisionTreeRegressor(random_state=571, min_samples_leaf=10)
).fit(X, y)

plt.figure(figsize=(12, 6))
plot_tree(sales_tree.named_steps["decisiontreeregressor"], max_depth=3)
plt.show()
```

:::

The tree should be read as a set of local comparisons. A split near the
root affects many observations; a split near a leaf affects a smaller
subgroup.

## Stopping and Pruning

A fully grown tree can overfit. Tree complexity can be controlled by
minimum node size, maximum depth, minimum improvement, or
cost-complexity pruning.

Cost-complexity pruning chooses a subtree by minimizing

$$
\sum_{m=1}^{|T|} \sum_{i:x_i \in R_m}(y_i-\hat y_{R_m})^2
+ \alpha |T|.
$$

The tuning parameter $\alpha$ penalizes tree size. Larger values produce
smaller trees.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
best_cp <- sales_tree$cptable[
  which.min(sales_tree$cptable[, "xerror"]),
  "CP"
]
pruned_tree <- prune(sales_tree, cp = best_cp)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
tree_model = sales_tree.named_steps["decisiontreeregressor"]
path = tree_model.cost_complexity_pruning_path(
    sales_tree.named_steps["columntransformer"].transform(X),
    y
)
ccp_alphas = path.ccp_alphas
print(ccp_alphas[:10])
```

:::

Cross-validation should be used to choose the pruning level. A smaller
tree with nearly the same error is often preferable because it is easier
to explain and less sensitive to noise.

## Strengths and Limits

Trees naturally capture:

- nonlinear relationships,
- interactions,
- threshold effects,
- mixed numeric and categorical predictors.

Their main weakness is variance. A small change in the training data
can produce a different first split, which changes the whole tree. A
single tree is therefore useful for interpretation and exploration, but
not always the most accurate predictor.
