---
title: "Ensemble Methods"
---

Ensembles improve prediction by combining many models. A single tree
can have high variance; bagging, random forests, and boosting use many
trees to produce more accurate and stable predictions.

## Learning Objectives

- Explain why averaging many models can reduce variance.
- Compare bagging, random forests, and boosting.
- Tune ensemble models using resampling.
- Interpret variable importance and partial dependence with caution.
- Evaluate whether extra accuracy is worth lower transparency.

## Averaging and Variance Reduction

If several predictors are unbiased and not perfectly correlated,
averaging them can reduce variance. Bagging applies this idea to trees:
fit many trees on bootstrap samples and average their predictions.

For regression,

$$
\hat f_{\text{bag}}(x)
= \frac{1}{B}\sum_{b=1}^{B}\hat f^{*b}(x).
$$

For classification, trees can vote or average class probabilities.

## Bagging

Bagging uses bootstrap samples to create many training sets. Each tree
is usually grown deeply, so individual trees are flexible and unstable.
The average is more stable than any one tree.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(randomForest)

bag_fit <- randomForest(
  High ~ .,
  data = train,
  mtry = ncol(train) - 1,
  importance = TRUE
)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.ensemble import RandomForestClassifier

bag_fit = RandomForestClassifier(
    n_estimators=500,
    max_features=None,
    random_state=571
)
bag_fit.fit(X_train, y_train)
```

:::

When `mtry` equals the number of predictors, each split can consider
all variables. This is bagging for tree models.

## Random Forests

Random forests modify bagging by allowing each split to consider only a
random subset of predictors. This decorrelates the trees. If one
predictor is very strong, ordinary bagging may use it near the top of
many trees, making the trees similar. Random predictor selection gives
other variables a chance to define different partitions.

Important tuning parameters include:

| Parameter | Effect |
|---|---|
| Number of trees | More trees stabilize the average |
| `mtry` | Smaller values decorrelate trees more strongly |
| Node size | Controls depth and local flexibility |
| Sampling scheme | Affects class balance and bootstrap variation |

## Boosting

Boosting builds an additive model sequentially:

$$
\hat f(x) = \sum_{b=1}^{B}\lambda \hat f_b(x).
$$

Each new tree focuses on structure that the current ensemble has not
captured well. The learning rate $\lambda$ controls how much each tree
contributes. Smaller learning rates usually require more trees but can
generalize better.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(gbm)

boost_fit <- gbm(
  Sales ~ .,
  data = train,
  distribution = "gaussian",
  n.trees = 3000,
  interaction.depth = 3,
  shrinkage = 0.01,
  cv.folds = 5
)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.ensemble import GradientBoostingRegressor

boost_fit = GradientBoostingRegressor(
    n_estimators=3000,
    learning_rate=0.01,
    max_depth=3,
    random_state=571
)
boost_fit.fit(X_train, y_train)
```

:::

Boosting is powerful, but it has more tuning choices: number of trees,
learning rate, interaction depth, and minimum observations in terminal
nodes.

## Comparing Tree Ensembles

A disciplined comparison uses the same split or resampling plan for all
models:

1. fit a pruned single tree,
2. fit a bagged tree ensemble,
3. fit a random forest,
4. fit a boosted tree model,
5. tune only inside training folds,
6. evaluate the selected models on held-out data.

The comparison should include both predictive performance and the cost
of complexity.

## Interpretation

Ensembles are less transparent than single trees, but several summaries
help:

| Tool | Use | Caution |
|---|---|---|
| Variable importance | Ranks influential predictors | Can be biased by scale or cardinality |
| Partial dependence | Shows average fitted effect | Can be misleading with strong interactions |
| Individual conditional curves | Shows heterogeneity across observations | Can be crowded and noisy |
| Test performance | Measures generalization | Does not explain the model alone |

Interpretation should be tied to the fitted model and the data
distribution, not treated as a causal explanation.
