---
title: "Polynomial Regression and Step Functions"
---

Polynomial regression and step functions are two direct ways to move
beyond a straight-line relationship while staying inside a regression
framework. Polynomials create smooth global curves; step functions
create piecewise constant regions.

## Learning Objectives

- Fit polynomial regression models.
- Explain the difference between raw and orthogonal polynomial terms.
- Fit step functions with cut points.
- Use validation to choose polynomial degree or number of bins.
- Communicate nonlinear fitted relationships with prediction plots.

## Polynomial Terms

A degree-$d$ polynomial regression has the form

$$
\operatorname{E}(Y \mid X=x)
= \beta_0 + \beta_1x + \beta_2x^2 + \cdots + \beta_dx^d.
$$

The model is linear in the coefficients even though it is nonlinear in
$x$.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
set.seed(571)
n <- 250
nonlinear <- data.frame(age = runif(n, 18, 70))
nonlinear$wage <- 18 + 2.1 * nonlinear$age -
  0.025 * nonlinear$age^2 + rnorm(n, sd = 8)

poly2 <- lm(wage ~ poly(age, 2), data = nonlinear)
poly4 <- lm(wage ~ poly(age, 4), data = nonlinear)
summary(poly2)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
import pandas as pd
from sklearn.linear_model import LinearRegression
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import PolynomialFeatures

rng = np.random.default_rng(571)
n = 250
nonlinear = pd.DataFrame({"age": rng.uniform(18, 70, n)})
nonlinear["wage"] = (
    18 + 2.1 * nonlinear["age"] -
    0.025 * nonlinear["age"] ** 2 +
    rng.normal(0, 8, n)
)

poly2 = make_pipeline(PolynomialFeatures(2, include_bias=False), LinearRegression())
poly4 = make_pipeline(PolynomialFeatures(4, include_bias=False), LinearRegression())
poly2.fit(nonlinear[["age"]], nonlinear["wage"])
poly4.fit(nonlinear[["age"]], nonlinear["wage"])
```

:::

High-degree polynomials can behave poorly near the boundaries of the
data. Validation should decide whether additional degree earns better
generalization.

## Step Functions

A step function turns a continuous predictor into intervals:

$$
\operatorname{E}(Y \mid X)
= \beta_0 + \beta_1I(X \in C_1) + \cdots + \beta_KI(X \in C_K).
$$

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
step_fit <- lm(wage ~ cut(age, breaks = 5), data = nonlinear)
summary(step_fit)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import statsmodels.formula.api as smf

step_data = nonlinear.copy()
step_data["age_bin"] = pd.cut(step_data["age"], bins=5)
step_fit = smf.ols("wage ~ C(age_bin)", data=step_data).fit()
print(step_fit.summary())
```

:::

Step functions are easy to explain but can create artificial jumps at
cut points. They are useful when thresholds are substantively
meaningful or when interpretability matters more than smoothness.
