---
title: "Support Vector Machines"
---

Support vector machines use margins and kernels to construct
classifiers that can be linear in a transformed feature space while
nonlinear in the original variables.

## Learning Objectives

- Explain maximal-margin and soft-margin classification.
- Interpret support vectors and the cost parameter.
- Compare linear, polynomial, and radial kernels.
- Explain why scaling is essential for SVMs.
- Tune cost and kernel parameters with cross-validation.

## Separating Hyperplanes

For classes coded as $y_i \in \{-1,1\}$, a separating hyperplane has
the form

$$
f(x) = \beta_0 + x^\top\beta.
$$

The predicted class is the sign of $f(x)$. If the classes are perfectly
separable, many hyperplanes may classify the training data correctly.
The maximal-margin classifier chooses the hyperplane with the largest
distance to the closest training points.

Those closest points are the support vectors. They determine the
boundary; observations far from the boundary do not directly control
the fit.

## Soft Margins

Perfect separation is rare and can lead to overly sensitive boundaries.
The soft-margin classifier allows violations:

$$
y_i(\beta_0 + x_i^\top\beta) \geq 1 - \xi_i,
\qquad \xi_i \geq 0.
$$

The cost parameter controls the penalty for violations. A large cost
allows fewer violations and may produce a narrower margin. A small cost
allows more violations and may produce a wider, more stable margin.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(e1071)

linear_svm <- svm(
  class ~ .,
  data = train,
  kernel = "linear",
  cost = 1,
  scale = TRUE
)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import StandardScaler
from sklearn.svm import SVC

linear_svm = make_pipeline(
    StandardScaler(),
    SVC(kernel="linear", C=1)
)
linear_svm.fit(X_train, y_train)
```

:::

## Kernels

Kernels compute similarity after an implicit feature transformation.
This allows nonlinear boundaries without explicitly creating all
transformed variables.

| Kernel | Form | Boundary behavior |
|---|---|---|
| Linear | $x_i^\top x_j$ | Linear in original predictors |
| Polynomial | $(c + x_i^\top x_j)^d$ | Global curved boundaries |
| Radial basis | $\exp(-\gamma\|x_i-x_j\|^2)$ | Local flexible boundaries |

The radial kernel is often effective because it can create local
nonlinear boundaries. Its flexibility depends on $\gamma$: larger
values create more local influence.

## Iris Decision Surfaces

A two-feature iris classification example makes kernels visible. Fit
linear, polynomial, and radial SVMs on the same scaled predictors, then
plot the predicted class across a grid. The linear kernel creates a
straight boundary in the two-dimensional plot. The polynomial and
radial kernels can bend around class structure.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
tuned <- tune(
  svm,
  class ~ .,
  data = train,
  kernel = "radial",
  ranges = list(cost = c(0.1, 1, 10), gamma = c(0.1, 1, 5))
)

best_svm <- tuned$best.model
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.model_selection import GridSearchCV
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import StandardScaler
from sklearn.svm import SVC

svm_pipeline = make_pipeline(StandardScaler(), SVC(kernel="rbf"))
tuned = GridSearchCV(
    svm_pipeline,
    param_grid={
        "svc__C": [0.1, 1, 10],
        "svc__gamma": [0.1, 1, 5]
    },
    cv=5
)
tuned.fit(X_train, y_train)
best_svm = tuned.best_estimator_
```

:::

This comparison should emphasize that better visual separation on the
training data is not enough. Kernel choice and tuning parameters must
be chosen with validation or cross-validation.

## Scaling and Tuning

SVMs depend on distances and inner products, so feature scale matters.
If one predictor has a much larger numeric range than another, it can
dominate the margin and kernel values.

The usual tuning sequence is:

1. split the data,
2. center and scale predictors inside the training data,
3. choose candidate kernels,
4. tune cost and kernel-specific parameters by cross-validation,
5. evaluate the selected model on held-out data,
6. inspect support vectors and decision surfaces where possible.
