---
title: "Splines and Generalized Additive Models"
---

Splines and generalized additive models provide flexible nonlinear
regression without forcing one global polynomial to describe the entire
relationship. They are especially useful when the fitted curve should
be smooth but locally adaptive.

## Learning Objectives

- Explain knots, basis functions, and degrees of freedom.
- Fit regression splines and natural splines.
- Describe generalized additive models.
- Use validation to choose smoothness.
- Interpret fitted smooth terms with prediction plots.

## Regression Splines

A regression spline fits piecewise polynomials that join at knots. The
number and placement of knots control flexibility.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(splines)

set.seed(571)
n <- 250
spline_data <- data.frame(age = runif(n, 18, 70))
spline_data$wage <- 20 + 1.8 * spline_data$age -
  0.02 * spline_data$age^2 + sin(spline_data$age / 5) * 4 +
  rnorm(n, sd = 7)

spline_fit <- lm(wage ~ bs(age, df = 6), data = spline_data)
natural_fit <- lm(wage ~ ns(age, df = 5), data = spline_data)
summary(spline_fit)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
import pandas as pd
import statsmodels.formula.api as smf

rng = np.random.default_rng(571)
n = 250
spline_data = pd.DataFrame({"age": rng.uniform(18, 70, n)})
spline_data["wage"] = (
    20 + 1.8 * spline_data["age"] -
    0.02 * spline_data["age"] ** 2 +
    np.sin(spline_data["age"] / 5) * 4 +
    rng.normal(0, 7, n)
)

spline_fit = smf.ols("wage ~ bs(age, df=6)", data=spline_data).fit()
natural_fit = smf.ols("wage ~ cr(age, df=5)", data=spline_data).fit()
print(spline_fit.summary())
```

:::

Natural splines add boundary constraints, which often stabilize fitted
curves near the smallest and largest observed values.

## Generalized Additive Models

A generalized additive model has the form

$$
g\{\operatorname{E}(Y \mid X)\}
= \beta_0 + f_1(X_1) + f_2(X_2) + \cdots + f_p(X_p).
$$

The model is additive across predictors but nonlinear within each
predictor.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
library(gam)

gam_fit <- gam(wage ~ s(age, df = 5), data = spline_data)
summary(gam_fit)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
gam_like_fit = smf.ols("wage ~ bs(age, df=5)", data=spline_data).fit()
print(gam_like_fit.summary())
```

:::

The fitted smooth term should be interpreted with predicted values and
uncertainty, not as a single slope.
