---
title: "Unsupervised Learning"
---

Unsupervised learning searches for structure without a designated
response variable. Because there is no outcome to predict, the quality
of the analysis depends on preprocessing, distance measures, stability,
and whether the discovered structure answers a meaningful question.

## Learning Objectives

- Distinguish supervised prediction from unsupervised structure
  discovery.
- Use PCA for visualization and compression.
- Compare k-means and hierarchical clustering.
- Explain how scaling and distance metrics affect clusters.
- Evaluate unsupervised results with stability and interpretability.

## No Response, No Single Test Error

In supervised learning, held-out error provides a natural evaluation
target. In unsupervised learning, there is no response variable, so
there is no single test error to optimize. Instead, analysts ask:

- Are the patterns stable under reasonable preprocessing choices?
- Are clusters separated enough to describe?
- Do the patterns persist across resamples?
- Are the discovered groups useful for a later scientific or business
  question?
- Could the result be an artifact of scale, outliers, or distance
  choice?

## PCA for Structure

PCA is often used to visualize high-dimensional data before clustering.
It creates uncorrelated component scores ordered by explained variance.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
x <- scale(USArrests)
pca <- prcomp(x)
plot(pca$x[, 1], pca$x[, 2], pch = 19)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.decomposition import PCA
from sklearn.preprocessing import StandardScaler
import matplotlib.pyplot as plt

x = StandardScaler().fit_transform(us_arrests)
pca = PCA().fit_transform(x)

plt.scatter(pca[:, 0], pca[:, 1])
plt.xlabel("PC1")
plt.ylabel("PC2")
plt.show()
```

:::

The first two components may reveal broad separation among
observations, but they do not guarantee meaningful clusters. PCA finds
directions of high variance, not necessarily directions that match a
substantive grouping.

## K-Means Clustering

K-means partitions observations into $K$ clusters by minimizing
within-cluster squared distance:

$$
\sum_{k=1}^{K}\sum_{i\in C_k}\|x_i-\bar x_k\|^2.
$$

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
set.seed(571)
km3 <- kmeans(x, centers = 3, nstart = 20)
km3$cluster
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from sklearn.cluster import KMeans

km3 = KMeans(n_clusters=3, n_init=20, random_state=571).fit(x)
print(km3.labels_)
```

:::

The analyst chooses $K$. Common aids include within-cluster sum of
squares, silhouette summaries, stability checks, and interpretability
of cluster profiles. None of these automatically proves the "true"
number of clusters.

## Hierarchical Clustering

Hierarchical clustering builds a tree of nested groupings. It requires
a distance metric and linkage rule.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
hc_complete <- hclust(dist(x), method = "complete")
plot(hc_complete)
rect.hclust(hc_complete, k = 3)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
from scipy.cluster.hierarchy import dendrogram, fcluster, linkage
import matplotlib.pyplot as plt

hc_complete = linkage(x, method="complete")
dendrogram(hc_complete)
plt.show()
clusters = fcluster(hc_complete, t=3, criterion="maxclust")
```

:::

Complete linkage tends to create compact clusters. Single linkage can
chain observations together. Average linkage is often a middle ground.
Changing linkage can change the dendrogram, so the choice should be
reported.

## Scaling and Distance

Clustering is often dominated by scale. If one variable has a much
larger numeric range, Euclidean distance will emphasize that variable.
Standardization is common, but it is not automatic truth. Scaling says
that one standard deviation in each variable should count equally.

Outliers also matter. A small number of unusual observations can affect
cluster centers, dendrogram structure, and PCA directions. The opening
analysis should therefore include summaries, plots, and sensitivity
checks.

## Cluster Profiles

Clusters become useful when they can be described. After assigning
clusters, summarize original variables by cluster:

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
aggregate(USArrests, by = list(cluster = km3$cluster), mean)
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
profiles = us_arrests.assign(cluster=km3.labels_).groupby("cluster").mean()
print(profiles)
```

:::

Profiles should be written in the original measurement units when
possible. A cluster label is not an explanation; it is a concise name
for a pattern that must be interpreted in context.
