---
title: "Applied Statistical Learning Case Studies"
---

Applied statistical learning often combines many course tools in one
analysis: exploratory summaries, transformations, probability models,
simulation, model comparison, and careful communication. This chapter
uses scientific and policy-style examples to show how the pieces fit
together.

## Learning Objectives

- Translate a scientific or policy question into a modeling workflow.
- Combine distribution fitting, regression, and uncertainty summaries.
- Explain how simulation supports decision-making.
- Report results without overstating causal or predictive claims.
- Identify where model assumptions matter for a final recommendation.

## Size Limits and Reproductive Contribution

A fishery management question might ask whether smaller mature fish
contribute meaningfully to local replenishment, even when policy rules
focus on larger fish. The statistical workflow is not a single model.
It includes data cleaning, length conversion, distribution summaries,
assignment probabilities, and uncertainty.

Key modeling tasks include:

1. define the population and sampling frame,
2. convert measurements onto a common length scale,
3. compare size distributions across management zones,
4. estimate reproductive contribution by size class,
5. summarize uncertainty with confidence intervals or simulation,
6. translate the result into a policy-relevant statement.

## Distribution-Fitting Frame

Distribution fitting is useful when the goal is to describe a size,
waiting-time, abundance, or loss distribution.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
set.seed(571)
fish <- data.frame(
  length_cm = rgamma(500, shape = 9, scale = 4.2),
  zone = sample(c("fished", "reserve"), 500, replace = TRUE)
)

by(fish$length_cm, fish$zone, summary)
fit_gamma <- MASS::fitdistr(fish$length_cm, densfun = "gamma")
fit_gamma
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np
import pandas as pd
from scipy import stats

rng = np.random.default_rng(571)
fish = pd.DataFrame({
    "length_cm": rng.gamma(shape=9, scale=4.2, size=500),
    "zone": rng.choice(["fished", "reserve"], size=500)
})

print(fish.groupby("zone")["length_cm"].describe())
shape, loc, scale = stats.gamma.fit(fish["length_cm"], floc=0)
print({"shape": shape, "scale": scale})
```

:::

The fitted distribution should be checked against the observed data.
Goodness-of-fit summaries are evidence, not guarantees.

## Simulation for Decision Support

Simulation can propagate uncertainty from biological or operational
inputs into a policy-relevant quantity.

:::{.panel-tabset group="language"}

## R

```{r}
#| echo: true
#| eval: false
#| fig-align: center
set.seed(571)
simulate_contribution <- function(n, mean_small, mean_large) {
  small <- rgamma(n, shape = 25, rate = 25 / mean_small)
  large <- rgamma(n, shape = 25, rate = 25 / mean_large)
  small / (small + large)
}

share_small <- simulate_contribution(10000, mean_small = 36, mean_large = 64)
quantile(share_small, c(0.05, 0.5, 0.95))
```

## Python

```{python}
#| echo: true
#| eval: false
#| fig-align: center
import numpy as np

def simulate_contribution(n, mean_small, mean_large, seed=571):
    rng = np.random.default_rng(seed)
    small = rng.gamma(shape=25, scale=mean_small / 25, size=n)
    large = rng.gamma(shape=25, scale=mean_large / 25, size=n)
    return small / (small + large)

share_small = simulate_contribution(10000, mean_small=36, mean_large=64)
print(np.quantile(share_small, [0.05, 0.5, 0.95]))
```

:::

The simulation does not settle the policy question by itself. It makes
the consequences of assumptions visible and gives the analyst a way to
describe uncertainty in the decision quantity.
