---
title: "Lab 11: Support Vector Machines"
execute:
  enabled: false
  eval: false
  warning: false
  message: false
format:
  html:
    toc: true
    number-sections: true
---
 Support Packages

Please hide/remove the unnecessary support packages/functions. My `ggplot` theme parameters are removed. You can use them from the previous lab and assignment. Please remember the goal for giving you the theme skeleton is make you design yours. Although, **Beauty Lies in the Eye of the Beholder**.

```{r support, warning=FALSE,message=FALSE,include=TRUE}
require(MASS)
require(glmnet)
library(flextable)
library(leaps)
library(GGally)
library(caret)
library(readr)
library(reticulate)
library(corrplot)
library(ggfortify)
library(MASS)
library(zoo)
library(doParallel)
library(lubridate)
library(reshape2)
library(data.table)
library(klaR)
library(feather)
if (!require("pacman")) install.packages("pacman")
pacman::p_load("weights", "interactions", "cjoint", "plm", "interactions", "jtools", "stats", "miceadds", "broom", "RColorBrewer", "ggstatsplot", "ggpubr", "stargazer", "sandwich", "hrbrthemes", "rms", "interplot", "coefplot", "gmodels", "car", "lattice","foreign", "ggplot2", "MASS", "Hmisc", "reshape2", "oddsratio", "tidyr", "psych", "dplyr", "tidyverse", "cjoint", "ISLR","ISLR2","kableExtra","ggExtra","gridExtra")

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE,echo = TRUE)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

no_cores <- detectCores() - 4  
registerDoParallel(cores=no_cores)  
```

# Data Import and Cleaning and Exploration.

The purpose of this study is to examine whether commercial wearable devices (Apple Watch and Fitbit)
can accurately predict sedentary behavior and light, moderate, and vigorous physical activities and to
develop indicators to assist in predictions for future studies. It's an okay [paper](https://europepmc.org/article/ppr/ppr123211) to read, however the data is what we want. The methods and results are summarized below. We will only run SVM with different kernels ans compare the results.


## Methods

We recruited a convenience sample of 49 participants (23 men and 26 women) to wear three
devices, an Apple Watch Series 2, a Fitbit Charge HR2, and and iPhone 6S. Participants completed a 65-
minute protocol consisting of 40 minutes of total treadmill time and 25 minutes of sitting or lying time.
Indirect calorimetry was used to measure energy expenditure. The outcome variable for the study was the
activity class; lying, sitting, walking self-paced, and running 3 METs, 5 METs, and 7 METs. Minute-by-minute heart rate, steps, distance, and calories from Apple Watch and Fitbit were included in four different
machine learning models.

## Results

Our dataset included 3656 and 2608 minutes of Apple Watch and Fitbit data, respectively. We
tested decision trees, support vector machines, random forest, and rotation forest models. Rotation forest
models had the highest classification accuracies at 82.6% for Apple Watch and 89.3% for Fitbit.
Classification ac curacies for Apple Watch data ranged from 72.5% for sitting to 89.0% for 7 METs . For
Fitbit, accuracies varied between 86.2% for sitting to 92.6% for 7 METs.

```{r importdata, warning=FALSE,message=FALSE}
set.seed(123)
library(mlbench)
library(caret)
library(caretEnsemble)
library(e1071)
library(tidyverse)

fitbit <- data.frame(read.csv(file="data/aw_fb_data.csv")) # data for prediction
  
# fitbit <- fitbit %>%  mutate(activity_num = case_when(
#     activity == "Lying" ~1,
#     activity == "Running 3 METs" ~ 2,
#     activity == "Running 5 METs" ~ 3,
#     activity == "Running 7 METs" ~ 4,
#     activity == "Self Pace walk" ~ 5,
#     activity == "Sitting" ~ 6))
# fitbit$activity_num <- as.factor(fitbit$activity_num)
  
fitbit_clean <- fitbit[c(-1, -2)]

fitbit_clean %>% 
  mutate(activity = factor(activity, 
                        labels = make.names(levels(activity))))

my_index <- createDataPartition(fitbit_clean$activity, p = 0.70, list = F)
trainSet <- fitbit_clean[my_index, ]
testSet <- fitbit_clean[-my_index, ]

```

## Exploring the dataset (not needed). 

```{r eda, fig.align='center',fig.width=12, fig.height=12, dpi=300}
ggpairs(fitbit_clean, ggplot2::aes(colour = as.factor(activity), alpha = 0.4))

```


# Fitting Linear Kernal for SVM

## Fit the linear kernel to the train set with no scaling and cost parameter set to 10.

```{r svmlin, warning=FALSE,message=FALSE}
svm_lin <- svm(as.factor(activity) ~ ., data=trainSet,
                 kernel="linear",scale=F) #linear/radial/polynomial/sigmoid
```

## create the confusion matrix. 

Use [cvms](https://cran.r-project.org/web/packages/cvms/vignettes/Creating_a_confusion_matrix.html).

```{r confmatlin,fig.align='center',fig.height=5, fig.width=5, dpi=300, warning=FALSE, message=FALSE}
library(cvms)

conf_tab <- tibble("target" = as.factor(testSet$activity),
                  "prediction" = predict(svm_lin, testSet))

conf_mat <- confusion_matrix(targets = conf_tab$target,
                             predictions = conf_tab$prediction)

plot_confusion_matrix(conf_mat$`Confusion Matrix`[[1]])

```


# Fitting radial kernal for SVM

## Fit the radial kernel to the train set with no scaling and cost parameter set to 10.

```{r svmrad, warning=FALSE,message=FALSE}
svm_rad <- svm(as.factor(activity) ~ ., data=trainSet,
                 kernel="radial",scale=F) #linear/radial/polynomial/sigmoid
```

## create the confusion matrix. 

Use [cvms](https://cran.r-project.org/web/packages/cvms/vignettes/Creating_a_confusion_matrix.html).

```{r confmatrad,fig.align='center',fig.height=5, fig.width=5, dpi=300, warning=FALSE, message=FALSE}
library(cvms)

conf_tab <- tibble("target" = as.factor(testSet$activity),
                  "prediction" = predict(svm_rad, testSet))

conf_mat <- confusion_matrix(targets = conf_tab$target,
                             predictions = conf_tab$prediction)

plot_confusion_matrix(conf_mat$`Confusion Matrix`[[1]])

```



# Fitting polynomial kernal for SVM

## Fit the polynomial kernel to the train set with no scaling and cost parameter set to 10.

```{r svmpoly, warning=FALSE,message=FALSE}
svm_poly <- svm(as.factor(activity) ~ ., data=trainSet,
                 kernel="polynomial",scale=F, degree=4, cost=10) #linear/radial/polynomial/sigmoid
```

## create the confusion matrix. 

Use [cvms](https://cran.r-project.org/web/packages/cvms/vignettes/Creating_a_confusion_matrix.html).

```{r confmatpoly,fig.align='center',fig.height=5, fig.width=5, dpi=300, warning=FALSE, message=FALSE}
library(cvms)

conf_tab <- tibble("target" = as.factor(testSet$activity),
                  "prediction" = predict(svm_poly, testSet))

conf_mat <- confusion_matrix(targets = conf_tab$target,
                             predictions = conf_tab$prediction)

plot_confusion_matrix(conf_mat$`Confusion Matrix`[[1]])

```



# Fitting sigmoid kernal for SVM

## Fit the sigmoid kernel to the train set with no scaling and cost parameter set to 10.

```{r svmsig, warning=FALSE,message=FALSE}
svm_sig <- svm(as.factor(activity) ~ ., data=trainSet,
                 kernel="sigmoid",scale=F) #linear/radial/polynomial/sigmoid
```

## create the confusion matrix. 

Use [cvms](https://cran.r-project.org/web/packages/cvms/vignettes/Creating_a_confusion_matrix.html).

```{r confmatsig,fig.align='center',fig.height=5, fig.width=5, dpi=300, warning=FALSE, message=FALSE}
library(cvms)

conf_tab <- tibble("target" = as.factor(testSet$activity),
                  "prediction" = predict(svm_sig, testSet))

conf_mat <- confusion_matrix(targets = conf_tab$target,
                             predictions = conf_tab$prediction)

plot_confusion_matrix(conf_mat$`Confusion Matrix`[[1]])

```

# Compare the ROC Curves


```{r plotpredfunc, fig.align='center', fig.width=7,fig.height=4.5}

```


# Additional (not needed)


```{r}
# train_control <- trainControl(method="repeatedcv", number=10, repeats=3)

svm1 <- caret::train(as.factor(activity) ~., data = trainSet, method = "svmLinear",     trControl = trainControl(method = "repeatedcv", repeats = 5, 
classProbs =  TRUE))
#, trControl = train_control,  preProcess = c("center","scale"))

svm2 <- train(as.factor(activity) ~., data = trainSet, method = "svmLinear", tuneGrid = expand.grid(C = seq(0, 2, length = 20))) #trControl = train_control,  preProcess = c("center","scale"))

svm3 <- train(as.factor(activity) ~., data = trainSet, method = "svmRadial")# , trControl = train_control, preProcess = c("center","scale"), tuneLength = 10)

svm4 <- train(as.factor(activity) ~., data = trainSet, method = "svmPoly") #, trControl = train_control, preProcess = c("center","scale"), tuneLength = 4)

svm4 <- train(as.factor(activity) ~., data = trainSet, method = "rbf") #, trControl = train_control, preProcess = c("center","scale"), tuneLength = 4)
```

```{r}
svm1.pred <- predict(svm1, data=testSet)
```


# Python companion

Support vector machines require tuning because the margin penalty and kernel parameters determine the shape of the decision boundary.

```python
import pandas as pd
from sklearn.metrics import classification_report
from sklearn.model_selection import GridSearchCV, train_test_split
from sklearn.pipeline import Pipeline
from sklearn.preprocessing import StandardScaler
from sklearn.svm import SVC

df = pd.read_csv("data/lab11.csv")
X = pd.get_dummies(df.drop(columns="y"), drop_first=True)
y = df["y"]
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.30, stratify=y, random_state=571)

pipe = Pipeline([
    ("scale", StandardScaler()),
    ("svm", SVC(probability=True)),
])

search = GridSearchCV(
    pipe,
    param_grid={
        "svm__kernel": ["linear", "rbf"],
        "svm__C": [0.1, 1, 10],
        "svm__gamma": ["scale", 0.01, 0.1, 1],
    },
    cv=10,
    scoring="accuracy",
)
search.fit(X_train, y_train)
print(search.best_params_)
print(classification_report(y_test, search.predict(X_test)))
```

